---
title: "Sites not matched to an ePaddock"
date: last-modified
format:
  html:
    self-contained: true
    theme: cosmo
    toc: true
execute:
  echo: false
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup
#| include: false

library(targets)
library(dplyr)
library(leaflet)
library(stringr)
library(flextable)
library(glue)

store_path <- if (dir.exists("_targets")) "_targets" else "../../_targets"
targets::tar_load(c(data_site_paddocks, nearby_unmatched_paddocks), store = store_path)

# Sites with no matching ePaddock polygon
unmatched <- data_site_paddocks |>
  dplyr::filter(is.na(paddock_id)) |>
  dplyr::arrange(region_name, farmer, site_name, subsite_name)
```

## Overview

**`r nrow(unmatched)`** of **`r nrow(data_site_paddocks)`** sites did not fall
within any ePaddock polygon. Common causes are:

- Sites on road verges, fence-lines, or paddock boundaries
- Paddocks too small or irregular to appear in ePaddocks v3
- GPS coordinates recorded at the paddock entrance rather than inside
- ePaddocks coverage gaps (e.g. some Eyre Peninsula and QLD regions)

```{r}
#| label: summary-by-state

if (nrow(unmatched) == 0) {
  cat("No unmatched sites.")
} else {
  unmatched |>
    dplyr::count(state, name = "n_sites") |>
    flextable::flextable() |>
    flextable::set_header_labels(state = "State", n_sites = "Unmatched sites") |>
    flextable::autofit() |>
    flextable::theme_vanilla()
}
```

## Map

ePaddock boundaries within 1 km of each unmatched site are shown in grey.
Click a marker for full site details; points are coloured by state.

```{r}
#| label: map

m <- leaflet::leaflet() |>
  leaflet::addProviderTiles(leaflet::providers$Esri.WorldImagery)

# ePaddock boundaries (pre-loaded by the targets pipeline)
if (nrow(nearby_unmatched_paddocks) > 0) {
  m <- m |>
    leaflet::addPolygons(
      data        = nearby_unmatched_paddocks,
      color       = "#555555",
      weight      = 1,
      opacity     = 0.8,
      fill        = TRUE,
      fillColor   = "#cccccc",
      fillOpacity = 0.25,
      label       = ~as.character(id),
      group       = "ePaddocks"
    ) |>
    leaflet::addLayersControl(
      overlayGroups = "ePaddocks",
      options       = leaflet::layersControlOptions(collapsed = FALSE)
    )
}

# Unmatched site points
if (nrow(unmatched) > 0) {
  states <- sort(unique(unmatched$state))
  pal    <- leaflet::colorFactor(palette = "Set1", domain = states)

  popups <- glue::glue_data(unmatched,
    "<b>{str_to_title(site)}</b> — {str_to_title(subsite)}<br>
     <b>Farmer:</b> {str_to_title(farmer)}<br>
     <b>Region:</b> {str_to_title(region)}, {state}<br>
     <b>Coords:</b> {round(longitude, 4)}, {round(latitude, 4)}"
  )

  m <- m |>
    leaflet::addCircleMarkers(
      data        = unmatched,
      lng         = ~longitude,
      lat         = ~latitude,
      radius      = 7,
      stroke      = TRUE,
      weight      = 2,
      color       = "white",
      fillColor   = ~pal(state),
      fillOpacity = 0.9,
      popup       = popups,
      label       = ~str_to_title(paste(site, "-", subsite))
    ) |>
    leaflet::addLegend(
      position = "bottomright",
      pal      = pal,
      values   = unmatched$state,
      title    = "State"
    )
}

m
```

## Full site list

```{r}
#| label: table

if (nrow(unmatched) == 0) {
  cat("No unmatched sites.")
} else {
  unmatched |>
    dplyr::select(state, region, farmer, site, subsite, longitude, latitude) |>
    dplyr::mutate(
      longitude = round(longitude, 4),
      latitude  = round(latitude,  4)
    ) |>
    flextable::flextable() |>
    flextable::set_header_labels(
      state     = "State",
      region    = "Region",
      farmer    = "Farmer",
      site      = "Site",
      subsite   = "Subsite",
      longitude = "Lon",
      latitude  = "Lat"
    ) |>
    flextable::autofit() |>
    flextable::theme_vanilla() |>
    flextable::fontsize(size = 9, part = "all")
}
```
