---
title: "Section B Covariates: Sources, Processing & Outcomes"
subtitle: "How each temporally-varying covariate is built, and why"
format:
  html:
    self-contained: true
    theme: cosmo
    fig-width: 8
    fig-asp: 0.55
    code-fold: true
    code-summary: "Show code"
toc: true
toc-depth: 3
execute:
  echo: false
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup
#| include: false

library(targets)
library(terra)
library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(gt)

## Handle both pipeline rendering (from project root) and standalone rendering
## (from inside quarto_reports/), same convention as raw_data_update.qmd.
store_path <- if (dir.exists("_targets")) "_targets" else "../_targets"
r_path     <- if (dir.exists("_targets")) "r" else "../r"

## raster_mean_series() (r/b_raster_mean_series.R) -- collapses a SpatRaster
## to a tidy one-row-per-layer mean series, nationally or at points, used
## throughout this report. raster_zone_series() (r/b_raster_zone_series.R)
## is its per-AE-zone counterpart -- shows genuine regional divergence as a
## time series (how did zone A's own trend differ from zone B's), which a
## national mean hides and a few spatial-snapshot maps only show at
## isolated moments, not as a trend. raster_map_df() (r/b_raster_map_df.R)
## is kept for the one covariate (min_temp) where the spatial pattern
## itself, not its trend, is the point -- see that section for why. All
## sourced directly (not tar_load()'d, since they're plain functions, not
## targets) so this report and the pipeline itself share one implementation
## instead of the qmd keeping its own local copy.
source(file.path(r_path, "b_raster_mean_series.R"))
source(file.path(r_path, "b_raster_zone_series.R"))
source(file.path(r_path, "b_raster_map_df.R"))

## aez_zone_info() (r/a_aez_zone_info.R) -- region_label groups this
## report's own AE-zone line charts by, so e.g. 18 zones don't all land in
## one unreadable facet -- see the zone-series figures below.
source(file.path(r_path, "a_aez_zone_info.R"))

## rain_raster_coarse/min_temp_raster deliberately not loaded here -- neither
## is ever used as an R object in this report (both only ever appeared as
## plain text inside the pipeline-reference table's own string literals),
## so loading them just adds dead weight (large multi-decade SpatRaster
## stacks) to every render.
tar_load(
  c(
    aez_adj,
    gpp_raster_coarse,
    gpp_rolling_raster_3, gpp_rolling_anomaly_raster_3,
    gpp_rolling_raster_6, gpp_rolling_anomaly_raster_6,
    gpp_rolling_raster_12, gpp_rolling_anomaly_raster_12,
    gpp_viirs_frozen_blocks, gpp_viirs_recent_blocks,
    rain_rolling_raster_3, rain_rolling_anomaly_raster_3,
    rain_rolling_raster_6, rain_rolling_anomaly_raster_6,
    rain_rolling_raster_12, rain_rolling_anomaly_raster_12,
    rain_seasonal_raster, rain_seasonal_anomaly_raster,
    rain_whiplash_source_anomaly, rain_whiplash_source_window,
    rain_whiplash_trough_to_peak, rain_whiplash_window,
    soil_moisture_raster_coarse, soil_moisture_anomaly_raster,
    soil_moisture_whiplash_source_anomaly, soil_moisture_whiplash_source_window,
    soil_moisture_whiplash_trough_to_peak, soil_moisture_whiplash_window,
    aus_shp,
    min_temp_seasonal_raster, min_temp_year_blocks,
    paddock_season_covs
  ),
  store = store_path
)

## Years covered by the two most recent (non-frozen) GPP blocks -- used
## below to flag which months of the final GPP series could contain
## near-real-time, non-gap-filled-sourced data. See gpp_processing.qmd for
## the full source/splice/correction methodology this figure is drawn from.
gpp_recent_years <- sort(unlist(gpp_viirs_recent_blocks))

theme_set(
  theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(), legend.position = "bottom")
)

## Zone lookup for every AE-zone line chart below (region_label groups zones
## into readable facets; join once here rather than in every figure chunk).
zone_regions <- aez_zone_info() |> dplyr::select(ae_zone, region_label)

## Hand-picked, not ggplot's default hue wheel: with 16 zones (18 minus
## WA Ord/Tas Grain, which raster_zone_series() already drops for having no
## data) spread across facets of up to 5, an evenly-spaced global hue wheel
## puts same-facet zones (e.g. all 5 "WA *" zones, adjacent in aez_zone_info()'s
## own table and so alphabetically/visually grouped into one facet) right
## next to each other in hue space -- confirmed live, they rendered as
## near-indistinguishable pinks. Assigned instead so each facet's own zones
## are pulled from visually distant parts of the palette.
zone_colours <- c(
  `WA Northern` = "#1b9e77", `WA Central` = "#d95f02", `WA Sandplain` = "#7570b3",
  `WA Eastern` = "#e7298a", `WA Mallee` = "#66a61e",
  `SA Midnorth-Lower Yorke Eyre` = "#e6ab02",
  `SA Vic Bordertown-Wimmera` = "#a6761d", `SA Vic Mallee` = "#666666",
  `Vic High Rainfall` = "#377eb8", `NSW Vic Slopes` = "#e41a1c", `NSW Central` = "#4daf4a",
  `NSW NE Qld SE` = "#984ea3", `NSW NW Qld SW` = "#ff7f00",
  `Qld Central` = "#a65628", `Qld Burdekin` = "#f781bf", `Qld Atherton` = "#999999"
)

## Shared window-length colour scale for GPP/rainfall's own consolidated
## level/anomaly comparison figures -- one consistent colour per window across
## both panels, so a reader doesn't have to re-learn the colour key panel to
## panel. Whiplash no longer belongs to this family (2026-08) -- its own
## dedicated 30-month monthly-grain source isn't one of gpp_rolling_raster_3/6/12's
## windows, so it gets its own dedicated figure and colour instead, see
## @sec-whiplash.
window_colours <- c(`3-month` = "#66c2a5", `6-month` = "#fc8d62", `12-month` = "#8da0cb", `Seasonal` = "#e78ac3")
whiplash_colour <- "#b3452c"

## Computed once here and reused everywhere below (the summary table, the
## GPP source prose and the raw-monthly figure) -- gpp_raster_coarse (not
## monthly_gpp_rast) specifically: same full 1982-present monthly record,
## already resampled onto PML's coarse 0.1 degree grid, so terra::global()'s
## spatial mean is cheap here instead of running it over 300+ layers at
## GPP's native 500m resolution.
gpp_monthly_df <- raster_mean_series(gpp_raster_coarse) |>
  dplyr::mutate(recent_block = lubridate::year(date) %in% gpp_recent_years)
gpp_latest_month <- format(max(gpp_monthly_df$date), "%B %Y")

## Read straight from the pipeline's own rain_whiplash_window/soil_moisture_whiplash_window targets
## (not hand-typed) so this text can never silently drift out of sync with the actual value in use
## -- see _targets.R's own comment on rain_whiplash_window for why these are real targets. GPP has
## no whiplash entry any more (tried, dropped 2026-08 -- see @sec-whiplash for why).
whiplash_window_text <- if (rain_whiplash_window == soil_moisture_whiplash_window) {
  paste0(rain_whiplash_window, " months")
} else {
  paste0(rain_whiplash_window, " months for rainfall, ", soil_moisture_whiplash_window, " for soil moisture")
}

## Whiplash's source anomaly and its own rolling-average window length, in months -- read from the
## pipeline's own compute_rolling_mean_raster()/compute_whiplash_raster() call arguments in
## _targets.R (not hand-typed), used in the summary table below so it can't silently drift out of
## sync the way a hardcoded "2 years" once did.
rain_whiplash_source_text <- paste0(rain_whiplash_window, "-month window of a ", rain_whiplash_source_window, "-month rolling anomaly")
soil_moisture_whiplash_source_text <- paste0(soil_moisture_whiplash_window, "-month window of a ", soil_moisture_whiplash_source_window, "-month rolling anomaly")
```

## Overview

This section of the pipeline (`_targets.R`, section B) attaches four temporally-varying covariates -- Gross Primary Productivity (GPP), rainfall, soil moisture and minimum temperature -- to every structured-survey paddock, at a **seasonal** time step (with December considered Summer of the following year). Each covariate is downloaded from its own raw data source, processed through its own pipeline, and given whichever mix of a shared treatment menu (rolling averages of one or more window lengths, a raw snapshot, a per-season mean, and a "whiplash" swing measure) actually matches how that variable behaves ecologically -- not a single one-size-fits-all approach.

The final output of section B is `paddock_season_covs`: one row per paddock x season, feeding directly into the forecasting model.

```{r}
#| label: paddock-season-covs-glimpse

## One representative column per covariate "family" (level + whichever
## treatment is newest), not every column paddock_season_covs actually has
## (17 covariate columns total -- see the "At a glance" table below for the
## full list) -- this is just a glimpse of the final shape, not a full dump.
paddock_season_covs |>
  dplyr::select(paddock_id, season_year_adj, gpp_rolling3, gpp_rolling6, gpp,
                rain_rolling12, rain_seasonal,
                rain_whiplash_trough_to_peak, soil_moisture, min_temp) |>
  head(5) |>
  gt::gt() |>
  gt::fmt_number(columns = where(is.numeric), decimals = 3) |>
  gt::tab_options(table.font.size = "0.75em")
```

### At a glance

```{r}
#| label: summary-table

mt_latest_year <- max(unlist(min_temp_year_blocks))
gpp_history     <- paste0("1982-", gpp_latest_month, " (near-real-time)")

summary_tbl <- tibble::tribble(
  ~Covariate,             ~Source,                                                    ~`Native resolution`, ~`History used`,     ~`Temporal treatment`,                                                    ~`Final column(s)`,
  "GPP (rolling 3)",      "MODIS + VIIRS gap-filled/non-gap-filled (AppEEARS) / PML-V2", "500 m / 0.1°",     gpp_history,          "Trailing 3-month rolling average (mathematically the same as a plain seasonal mean once trimmed to season-end months, so there's no separate non-rolling GPP version)", "gpp_rolling3, gpp_rolling3_anomaly",
  "GPP (rolling 6)",      "MODIS + VIIRS gap-filled/non-gap-filled (AppEEARS) / PML-V2", "500 m / 0.1°",     gpp_history,          "Trailing 6-month rolling average",                                       "gpp_rolling6, gpp_rolling6_anomaly",
  "GPP (rolling 12)",     "MODIS + VIIRS gap-filled/non-gap-filled (AppEEARS) / PML-V2", "500 m / 0.1°",     gpp_history,          "Trailing 12-month rolling average",                                      "gpp, gpp_anomaly",
  "Rainfall (rolling 3/6/12)", "SILO monthly rainfall",                              "0.05°",              "1982+",              "Trailing 3-, 6- and 12-month rolling average",                           "rain_rolling3, rain_rolling3_anomaly, rain_rolling6, rain_rolling6_anomaly, rain_rolling12, rain_rolling12_anomaly",
  "Rainfall (seasonal)",  "SILO monthly rainfall",                                    "0.05°",              "1982+",              "Sum within each of the 4 seasons",                                       "rain_seasonal, rain_seasonal_anomaly",
  "Rainfall (whiplash)",  "SILO monthly rainfall",                                    "0.05°",              "1982+",              paste0("How far the current value has climbed above its own trough, trailing ", rain_whiplash_source_text), "rain_whiplash_trough_to_peak",
  "Soil moisture",        "AWRA-L v7 (sm_pct)",                                       "~5 km",              "1982+",              "Raw current-season snapshot, no windowing",                              "soil_moisture, soil_moisture_anomaly",
  "Soil moisture (whiplash)", "AWRA-L v7 (sm_pct)",                                   "~5 km",              "1982+",              paste0("How far the current value has climbed above its own trough, trailing ", soil_moisture_whiplash_source_text), "soil_moisture_whiplash_trough_to_peak",
  "Min temperature",      "SILO daily min_temp",                                      "0.05°",              paste0("1980-", mt_latest_year, " (5-year blocks)"), "Raw mean within each of the 4 seasons",            "min_temp"
)

summary_tbl |>
  gt::gt() |>
  gt::tab_options(table.font.size = "0.78em") |>
  gt::cols_width(Covariate ~ px(150))
```

All four covariates are coarsened onto PML-V2's shared 0.1&deg; grid before any temporal processing, so GPP, rainfall and soil moisture are directly comparable pixel-for-pixel (min_temp stays on SILO's native 0.05&deg; grid, extracted at each paddock's own point instead).

## Why different temporal treatments?

GPP, rainfall and soil moisture now draw from the same shared treatment menu -- a trailing rolling average (any window length, via `_targets.R`'s `gpp_rolling_windows`/`rain_rolling_windows`), a non-rolling per-season total or mean, and a "whiplash" swing measure (below) -- rather than each having its own bespoke, one-off calculation. What differs between them is **which items off that menu actually make ecological sense**, not the mechanics of computing them:

- **GPP** reflects whatever a farmer actually planted (or didn't) -- a decision made by the farmer, not fixed by the calendar. It gets three trailing rolling averages instead of a separate non-rolling option: **12-month** (`gpp`) captures whatever crop was grown regardless of when it was sown; **6-month** (`gpp_rolling6`) and **3-month** (`gpp_rolling3`) are more responsive alternatives alongside it, not replacements. There's no separate "this season's own mean" version, unlike rainfall below -- once trimmed to season-end months, a 3-month rolling average covers exactly the same 3 calendar months as a plain seasonal mean, so it would just be the same number computed twice. GPP does **not** get a whiplash treatment -- tried and dropped (2026-08), see [What is "whiplash"?](#sec-whiplash) below for why.
- **Rainfall** isn't a farmer decision, so there's no ecological reason to prefer one window over another a priori -- it gets the identical rolling-3/rolling-6/rolling-12 treatment as GPP, plus a simpler **sum within each season** on its own, kept alongside the rolling versions to check how much they actually differ (see [How redundant are the rainfall versions?](#how-redundant-are-the-rainfall-versions) below), and its own **whiplash**. There's no fixed Apr-Oct calendar-window version any more -- that was a once-a-year value broadcast identically across all 4 seasons, unlike everything else here, and rain_seasonal/rain_rolling6/rain_rolling12 already cover the same "how was the growing season" signal at the model's own seasonal grain (see `r_not_in_use/b_build_seasonal_window_raster.R` if it's needed again).
- **Soil moisture** is already an integrated state variable -- the AWRA water-balance model has already folded rainfall, evapotranspiration and drainage history into one number, so rolling or seasonally-summing it further would double-count memory that's already built in. It's used as-is: **whatever the most recent value is**, plus its own anomaly. It also gets a **whiplash** version, built the same way as GPP/rainfall's (@sec-whiplash) despite the "double-counting" concern above -- flagged as a live tension in `_targets.R`'s own comment on `soil_moisture_whiplash_source_raster`, not a resolved design choice.
- **Min temperature** represents a fixed physiological breeding-season switch (mice breed once winter cold eases), not a relative-to-local-norms effect -- so unlike the other three, it's **never converted to an anomaly**, and gets none of the rolling/seasonal/whiplash menu at all. Standardising or windowing it would risk masking the actual biological threshold that matters, regardless of what's "normal" for that particular paddock or how it got there.

```{r}
#| label: fig-lag-comparison
#| fig-cap: "How each covariate's value at a given season is actually built from raw monthly/daily data. GPP and rainfall's rolling windows slide forward every season; soil moisture and min_temp use only that season's own data, with no history carried in. Whiplash isn't shown here -- see 'What is whiplash?' below -- since it measures a swing across a window rather than an average within one."
#| fig-asp: 0.5

gpp12_row <- tibble::tibble(covariate = "GPP\n(rolling 12-month avg)", months_back = 0:11)
gpp6_row  <- tibble::tibble(covariate = "GPP\n(rolling 6-month avg)", months_back = 0:5)
rain12_row <- tibble::tibble(covariate = "Rainfall\n(rolling 12-month avg)", months_back = 0:11)
rain6_row  <- tibble::tibble(covariate = "Rainfall\n(rolling 6-month avg)", months_back = 0:5)
sm_row <- tibble::tibble(covariate = "Soil moisture\n(current snapshot)", months_back = 0)
mt_row <- tibble::tibble(covariate = "Min temp\n(that season only)", months_back = 0:2) # that season's 3 months

diagram_data <- dplyr::bind_rows(gpp12_row, gpp6_row, rain12_row, rain6_row, sm_row, mt_row) |>
  dplyr::mutate(covariate = factor(covariate, levels = rev(unique(covariate))))

ggplot(diagram_data, aes(x = -months_back, y = covariate, fill = covariate)) +
  geom_tile(height = 0.6, show.legend = FALSE) +
  scale_x_continuous(breaks = seq(-12, 0, 2), name = "Months before the season-end (0 = the season being scored)") +
  labs(y = NULL) +
  scale_fill_brewer(palette = "Set2")
```

### What is "whiplash"? {#sec-whiplash}

Whiplash targets a specific pattern: a **sustained** low period switching to a **sustained** high one -- not a fast month-to-month wobble. **Brown (2026, *Integrative Zoology*, `relevant_papers/`)** analysed 125 years of rainfall against confirmed Australian mouse-plague dates and found a plague requires a dry-phase deficit sustained 25-31 months, immediately followed by a wet-phase surplus sustained 17-20 months -- anything shorter on either end did not produce a plague in their data. Whiplash (`r/b_compute_whiplash_raster.R`) is this pipeline's own covariate for that same pattern: a rolling-window generalisation of Swain et al. (2018)'s year-over-year "precipitation whiplash" definition. Only the trough-to-peak (rising) direction is used -- a rapid recovery/increase matters for early warning of increasing mouse activity; `compute_whiplash_raster()` itself also supports the mirror-image `peak_to_trough` direction if a future use needs it.

**Source is deliberately its own dedicated rolling anomaly** (`rain_whiplash_source_anomaly`/`soil_moisture_whiplash_source_anomaly`), built at monthly grain and entirely separate from the `gpp_rolling_raster_3/6/12`/`rain_rolling_raster_3/6/12` family compared above. `compute_whiplash_raster()` then sweeps a **`r whiplash_window_text`** window on top of that slow source: at every point, it tracks the lowest value seen earlier in the window and reports how far *the current value* has climbed above that trough (`anchor = "current"`) -- not the biggest such recovery found anywhere in the window (`anchor = "anywhere"`, `compute_whiplash_raster()`'s other option), which can keep echoing an already-faded peak for most of the window's own length afterward.

```{r}
#| label: fig-whiplash-illustration
#| fig-cap: "How whiplash is measured (an illustrative synthetic example, not real pipeline data): at each point, the algorithm tracks the lowest value seen earlier in the window (grey dashed line). The solid arrow is what's actually reported for the current (rightmost) period -- anchor=\"current\", how far *now* has climbed above that trough. The faded dashed arrow shows what anchor=\"anywhere\" (compute_whiplash_raster()'s other option, not used in production) would report instead -- the single biggest such recovery anywhere in the window, which is still echoing an earlier, since-faded peak."
#| fig-asp: 0.45

## Reproduces compute_whiplash_raster()'s own running-min sweep
## (r/b_compute_whiplash_raster.R), on a plain illustrative vector instead of
## a SpatRaster -- not a call to the real pipeline function itself. Chosen so
## the current (last) period's own swing is genuinely smaller than the
## historical best -- the point being illustrated is that they differ.
example_anomaly <- tibble::tibble(
  period  = 1:16,
  anomaly = c(0.3, -0.5, -1.8, -0.9, 0.4, 1.6, 0.8, -0.2, -1.1, -0.3, 0.9, 2.1, 1.0, -0.6, -1.4, 0.5)
) |>
  dplyr::mutate(running_min = cummin(anomaly), swing = anomaly - running_min)

anywhere_point <- example_anomaly[which.max(example_anomaly$swing), ]
anywhere_trough <- example_anomaly |>
  dplyr::filter(period <= anywhere_point$period, anomaly == anywhere_point$running_min) |>
  dplyr::slice(1)

current_point  <- example_anomaly[nrow(example_anomaly), ]
current_trough <- example_anomaly |>
  dplyr::filter(period <= current_point$period, anomaly == current_point$running_min) |>
  dplyr::slice(1)

ggplot(example_anomaly, aes(period, anomaly)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_step(aes(y = running_min), colour = "grey50", linetype = "dashed") +
  geom_line(colour = "grey30") +
  geom_point(size = 1.8) +
  geom_segment(
    x = anywhere_trough$period, xend = anywhere_point$period,
    y = anywhere_trough$anomaly, yend = anywhere_point$anomaly,
    arrow = arrow(length = unit(0.16, "cm")), colour = "grey55", linetype = "22", linewidth = 0.6
  ) +
  annotate(
    "text", x = anywhere_point$period, y = anywhere_point$anomaly,
    label = paste0("anywhere = ", round(anywhere_point$swing, 1)), vjust = -0.9, hjust = 0.8, colour = "grey45", size = 3
  ) +
  geom_segment(
    x = current_trough$period, xend = current_point$period,
    y = current_trough$anomaly, yend = current_point$anomaly,
    arrow = arrow(length = unit(0.2, "cm")), colour = whiplash_colour, linewidth = 0.9
  ) +
  annotate(
    "text", x = current_point$period, y = current_point$anomaly,
    label = paste0("current = ", round(current_point$swing, 1)), vjust = -0.8, hjust = 1.05, colour = whiplash_colour, size = 3.5, fontface = "bold"
  ) +
  labs(x = "Period (trailing window)", y = "Anomaly (z-score)")
```

`compute_whiplash_raster()`'s first argument is exactly "which series to measure whiplash against", so redirecting either to a different source or window later is a one-line change in `_targets.R`, not a redesign.

### Validating against real plague occurrence {#sec-whiplash-validation}

An earlier version of this covariate (24-month source, 3-month sweep) was chosen from a
covariate-internal test: does whiplash detect a switch relative to its own history, at real
sustained-dry-to-sustained-wet transitions (e.g. the 2010 end of the Millennium Drought)? It did --
ranking 91st-100th percentile of its own distribution at the switch moment, vs. only 57th-74th
percentile for the raw source anomaly at the same moment. That's real, but it's the wrong target: it
never checked whether whiplash actually predicts real plague occurrence.

Checked directly against `raw_data/plague_occurrence/yearly_plague_occurrence.csv` (declared plague
years, 4 states, 1900-2024), with the cleanest test found: whiplash in a state-year that's eligible
to onset -- not itself a plague year, and not within the 2-year post-plague refractory window
plagues essentially never break regardless of conditions -- predicting a **new** onset the following
year, the 24/3 construction scored AUC=0.54 (p=0.99, indistinguishable from chance). A full sweep
(source 6-48 months x sweep 2-24 months x both anchor modes) found rainfall and soil moisture both
converge on the same clear optimum -- **30-month source, 12-month sweep, `anchor = "current"`**
(currently in production) -- while **GPP found no significant configuration anywhere in that same
sweep** (best AUC=0.62, p=0.29) and was dropped rather than shipped with an arbitrarily-chosen,
unvalidated spec.

```{r}
#| label: whiplash-plague-validation
#| fig-cap: "Rainfall and soil moisture whiplash (production spec) against each state's own declared plague years (grey bands), matching Brown (2026)'s own Figure 4c/5 shading convention."
#| fig-asp: 1.3

raw_data_path <- if (dir.exists("_targets")) "raw_data" else "../raw_data"

plague_raw <- readr::read_csv(file.path(raw_data_path, "plague_occurrence/yearly_plague_occurrence.csv"), show_col_types = FALSE)
names(plague_raw)[1] <- "Year"
plague_long <- plague_raw |>
  tidyr::pivot_longer(-Year, names_to = "state", values_to = "plague_flag") |>
  dplyr::mutate(plague = !is.na(plague_flag)) |>
  dplyr::select(year = Year, state, plague) |>
  dplyr::filter(year >= 1983, year <= 2024)

state_abbrev_to_full <- c(NSW = "New South Wales", QLD = "Queensland", SA = "South Australia", VIC = "Victoria")
aus_shp_4 <- aus_shp |> dplyr::filter(ST_NAME %in% names(state_abbrev_to_full)) |>
  dplyr::mutate(ST_NAME = unname(state_abbrev_to_full[ST_NAME]))

## Reuses raster_zone_series()'s own approach (r/b_raster_zone_series.R), grouped by aus_shp's real
## state polygons instead of aez_adj's AE zones -- yearly_plague_occurrence.csv is recorded at state
## grain, not AE-zone grain.
state_monthly <- function(rast, label) {
  extracted <- terra::extract(rast, terra::vect(aus_shp_4), fun = "mean", na.rm = TRUE, ID = FALSE)
  dates <- raster_layer_dates(rast)
  tibble::tibble(
    date = rep(dates, each = nrow(extracted)), state = rep(aus_shp_4$ST_NAME, times = length(dates)),
    value = as.numeric(as.matrix(extracted)), covariate = label
  )
}
monthly_wh <- dplyr::bind_rows(
  state_monthly(rain_whiplash_trough_to_peak, "Rainfall whiplash"),
  state_monthly(soil_moisture_whiplash_trough_to_peak, "Soil moisture whiplash")
) |> dplyr::filter(!is.na(value))

plague_years_marked <- plague_long |> dplyr::filter(plague)

ggplot(monthly_wh, aes(date, value, colour = covariate)) +
  geom_rect(
    data = plague_years_marked,
    aes(xmin = as.Date(paste0(year, "-01-01")), xmax = as.Date(paste0(year, "-12-31")), ymin = -Inf, ymax = Inf),
    inherit.aes = FALSE, fill = "grey70", alpha = 0.5
  ) +
  geom_line(linewidth = 0.4) +
  facet_wrap(~state, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c(`Rainfall whiplash` = "#2166ac", `Soil moisture whiplash` = "#c8722f")) +
  labs(x = NULL, y = "Whiplash (z-score units)", colour = NULL) +
  theme(legend.position = "top")
```

```{r}
#| label: whiplash-plague-validation-table

## Clean onset-forecasting test: whiplash in an eligible (non-plague, non-refractory) year
## predicting a NEW onset the following year -- see _targets.R's own comment on
## rain_whiplash_source_window for the full methodology.
plague_years_by_state <- plague_long |> dplyr::filter(plague) |> dplyr::select(state, plague_year = year)
all_state_years <- tidyr::expand_grid(state = unique(aus_shp_4$ST_NAME), year = 1983:2024)
recent_plague <- all_state_years |>
  dplyr::left_join(plague_years_by_state, by = "state", relationship = "many-to-many") |>
  dplyr::mutate(years_since = year - plague_year) |>
  dplyr::group_by(state, year) |>
  dplyr::summarise(in_refractory_window = any(years_since %in% 1:2, na.rm = TRUE), .groups = "drop")

evaluate_onset <- function(monthly_df, label) {
  annual <- monthly_df |>
    dplyr::mutate(year = as.integer(format(date, "%Y"))) |>
    dplyr::group_by(state, year) |>
    dplyr::summarise(value_max = max(value, na.rm = TRUE), .groups = "drop")
  onset_test <- annual |>
    dplyr::inner_join(plague_long, by = c("state", "year")) |> dplyr::filter(!plague) |>
    dplyr::inner_join(recent_plague, by = c("state", "year")) |> dplyr::filter(!in_refractory_window) |>
    dplyr::mutate(match_year = year + 1) |>
    dplyr::inner_join(plague_long, by = c("state", "match_year" = "year"), suffix = c("_this", "_next")) |>
    dplyr::filter(!is.na(value_max))
  yes <- onset_test$value_max[onset_test$plague_next]
  no  <- onset_test$value_max[!onset_test$plague_next]
  t <- t.test(yes, no)
  r <- rank(c(yes, no))
  auc <- (sum(r[seq_along(yes)]) - length(yes) * (length(yes) + 1) / 2) / (length(yes) * length(no))
  tibble::tibble(Covariate = label, `n (new onset)` = length(yes), `n (no onset)` = length(no),
                 `Mean before onset` = round(mean(yes), 3), `Mean before no onset` = round(mean(no), 3),
                 `p-value` = round(t$p.value, 4), AUC = round(auc, 3))
}

dplyr::bind_rows(
  evaluate_onset(monthly_wh |> dplyr::filter(covariate == "Rainfall whiplash"), "Rainfall whiplash"),
  evaluate_onset(monthly_wh |> dplyr::filter(covariate == "Soil moisture whiplash"), "Soil moisture whiplash")
) |>
  gt::gt() |>
  gt::tab_options(table.font.size = "0.82em")
```

The 30/12 pairing isn't a rejection of Brown's own 25-31-month dry-phase citation -- it's a
different point within the same validated range, picked by real outcome data instead of the
switch-event proxy. Two follow-up checks specifically on soil moisture: a matching sweep with **no**
rolling average at all (the raw monthly anomaly, testing whether the pipeline's own original
"AWRA already integrates history, don't smooth further" argument holds) topped out at AUC=0.61,
p=0.18 -- not significant, and meaningfully worse than the 30-month-rolled version. Shorter rolling
windows (3, 6 months) topped out similarly weak (AUC 0.59-0.61, all p>0.17). The rolling average is
doing real, validated work here, not redundant smoothing.

## GPP {#sec-gpp}

**Source:** MODIS + VIIRS (AppEEARS, gap-filled/non-gap-filled) / PML-V2 -- four raw products spliced into one continuous monthly record.
**Native resolution:** 500 m (coarsened to PML's shared 0.1&deg; grid before any temporal processing -- see the note below the "At a glance" table).
**Temporal span:** 1982 to `r gpp_latest_month` (the present month as of this report), near-real-time for the last month or two.

Four raw products are spliced into one continuous monthly record: PML-V2 (1982-1999), MODIS (2000-2011), VIIRS gap-filled (2012-present), with VIIRS non-gap-filled filling in the most recent months gap-filled hasn't published yet. Two of the three splices are bridged by a pixel-scale ratio correction; the third (MODIS into VIIRS) needs none, since the two products never share a date to compare and are treated as already continuous by design. **See [`gpp_processing.qmd`](gpp_processing.html) for the full source/processing/correction methodology and how well each splice holds up** -- this section just shows what the resulting record looks like once it feeds into the covariates the model actually consumes.

### The corrected GPP record

This is what the pipeline's final, corrected GPP record (`gpp_raster_coarse`, see `gpp_processing.qmd` for how it's spliced and corrected) actually looks like once it's rolled, seasonally summarised, and measured for whiplash.

```{r}
#| label: fig-gpp-raw-monthly
#| fig-cap: "Raw monthly GPP (national mean), not yet rolling-averaged -- every point is one calendar month's own value. Points in the last two years (red) may include months filled in from the bias-corrected non-gap-filled splice rather than gap-filled composites alone; everything before that is gap-filled-only and permanently frozen."
#| fig-asp: 0.4

ggplot(gpp_monthly_df, aes(date, value)) +
  geom_line(colour = "grey60", linewidth = 0.3) +
  geom_point(aes(colour = recent_block), size = 1.3) +
  scale_colour_manual(
    values = c(`FALSE` = "#1b7837", `TRUE` = "#e41a1c"),
    labels = c(`FALSE` = "Gap-filled, frozen", `TRUE` = "Recent block (gap-filled + possible near-real-time splice)"),
    name = NULL
  ) +
  labs(x = NULL, y = "GPP (gC/m²/day)")
```

### Comparing GPP's treatments {#sec-gpp-compare}

GPP's three rolling-window treatments -- level (top) and anomaly (bottom) -- coloured consistently by window, so a shorter window's extra responsiveness (and a longer window's extra smoothing) can be traced straight down from one panel to the other. GPP has no whiplash treatment (tried, dropped 2026-08 -- see @sec-whiplash-validation).

```{r}
#| label: fig-gpp-compare
#| fig-cap: "GPP's three rolling-window treatments (3/6/12-month), national mean, compared across level and anomaly."
#| fig-asp: 0.6

gpp_level_df <- dplyr::bind_rows(
  raster_mean_series(gpp_rolling_raster_3)  |> dplyr::mutate(window = "3-month"),
  raster_mean_series(gpp_rolling_raster_6)  |> dplyr::mutate(window = "6-month"),
  raster_mean_series(gpp_rolling_raster_12) |> dplyr::mutate(window = "12-month")
) |> dplyr::mutate(metric = "Level (gC/m²/day)")

gpp_anomaly_df <- dplyr::bind_rows(
  raster_mean_series(gpp_rolling_anomaly_raster_3)  |> dplyr::mutate(window = "3-month"),
  raster_mean_series(gpp_rolling_anomaly_raster_6)  |> dplyr::mutate(window = "6-month"),
  raster_mean_series(gpp_rolling_anomaly_raster_12) |> dplyr::mutate(window = "12-month")
) |> dplyr::mutate(metric = "Anomaly (z-score)")

gpp_compare_df <- dplyr::bind_rows(gpp_level_df, gpp_anomaly_df) |>
  dplyr::mutate(metric = factor(metric, levels = c("Level (gC/m²/day)", "Anomaly (z-score)")))

ggplot(gpp_compare_df, aes(date, value, colour = window)) +
  geom_hline(data = data.frame(metric = factor("Anomaly (z-score)", levels = levels(gpp_compare_df$metric))),
             aes(yintercept = 0), colour = "grey70", linewidth = 0.3, inherit.aes = FALSE) +
  geom_line() +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  scale_colour_manual(values = window_colours, name = "Rolling window") +
  labs(x = NULL, y = NULL)
```

### Where is this changing most? {#sec-gpp-zones}

A national mean can look unremarkable even when a real, sharp swing is happening in one specific region -- and a spatial map only shows a moment, not a trend. AE-zone lines show both at once: which zones are diverging from the national picture, and how their own trajectories differ from each other over time.

```{r}
#| label: fig-gpp-zone-anomaly
#| fig-cap: "GPP anomaly (12-month rolling, z-score) by AE zone, grouped into the same regions used elsewhere in this pipeline (r/a_aez_zone_info.R). Shared y-axis across facets -- anomaly is already standardised, so zones are directly comparable on it."
#| fig-asp: 1.1

gpp_zone_anomaly_df <- raster_zone_series(gpp_rolling_anomaly_raster_12, aez_adj) |>
  dplyr::left_join(zone_regions, by = "ae_zone")

ggplot(gpp_zone_anomaly_df, aes(date, value, colour = ae_zone)) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.3) +
  geom_line(linewidth = 0.35) +
  facet_wrap(~region_label, ncol = 2) +
  scale_colour_manual(values = zone_colours) +
  labs(x = NULL, y = "GPP anomaly (z-score)", colour = "AE zone") +
  theme(legend.position = "right", legend.text = element_text(size = 8), strip.text = element_text(size = 8.5))
```

## Rainfall {#sec-rainfall}

**Source:** SILO's pre-aggregated monthly rainfall totals (Bureau of Meteorology, 1889+), downloaded directly (no daily aggregation needed).
**Native resolution:** 0.05&deg; (resampled onto PML's coarser shared 0.1&deg; grid before any temporal processing).
**Temporal span:** 1982+ (trimmed from SILO's own full 1889+ history to match the other covariates' common start).

**Processing:** resample onto PML's 0.1&deg; grid &rarr; trim to 1982+ &rarr; **three parallel versions**: (1) trailing 3-, 6- and 12-month rolling averages (`tar_map()` over `rain_rolling_windows`, the identical mechanism GPP uses above), (2) summed within each of the 4 calendar seasons independently, plus (3) whiplash (@sec-whiplash). Each rolling/seasonal version gets its own leave-one-out z-score anomaly (purely per-pixel, no neighbour-blending -- SILO is already a smoothed national surface).

### Comparing rainfall's treatments {#sec-rain-compare}

The three rolling windows (3/6/12-month) compared across level and anomaly, the same layout as GPP's own comparison above. The per-season total is kept as its own figure below, not folded in here -- it's a **sum** over 3 months, not an average like the rolling versions, so overlaying it on the same axis would compare different units, not just different windows. Whiplash is built from an entirely separate 30-month monthly-grain source (@sec-whiplash), not derived from any of these three, so it gets its own figure too.

```{r}
#| label: fig-rain-compare
#| fig-cap: "Rainfall's rolling 3-, 6- and 12-month treatments, national mean, compared across level and anomaly."
#| fig-asp: 0.6

rain_level_df <- dplyr::bind_rows(
  raster_mean_series(rain_rolling_raster_3)  |> dplyr::mutate(window = "3-month"),
  raster_mean_series(rain_rolling_raster_6)  |> dplyr::mutate(window = "6-month"),
  raster_mean_series(rain_rolling_raster_12) |> dplyr::mutate(window = "12-month")
) |> dplyr::mutate(metric = "Level (mm/month)")

rain_anomaly_df <- dplyr::bind_rows(
  raster_mean_series(rain_rolling_anomaly_raster_3)  |> dplyr::mutate(window = "3-month"),
  raster_mean_series(rain_rolling_anomaly_raster_6)  |> dplyr::mutate(window = "6-month"),
  raster_mean_series(rain_rolling_anomaly_raster_12) |> dplyr::mutate(window = "12-month")
) |> dplyr::mutate(metric = "Anomaly (z-score)")

rain_compare_df <- dplyr::bind_rows(rain_level_df, rain_anomaly_df) |>
  dplyr::mutate(metric = factor(metric, levels = c("Level (mm/month)", "Anomaly (z-score)")))

ggplot(rain_compare_df, aes(date, value, colour = window)) +
  geom_hline(data = data.frame(metric = factor("Anomaly (z-score)", levels = levels(rain_compare_df$metric))),
             aes(yintercept = 0), colour = "grey70", linewidth = 0.3, inherit.aes = FALSE) +
  geom_line() +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  scale_colour_manual(values = window_colours, name = "Rolling window") +
  labs(x = NULL, y = NULL)
```

```{r}
#| label: fig-rain-whiplash
#| fig-cap: "Rainfall's whiplash source (its own dedicated 30-month rolling anomaly, top) and the resulting whiplash swing (bottom), national mean."
#| fig-asp: 0.6

rain_whiplash_source_df <- raster_mean_series(rain_whiplash_source_anomaly) |> dplyr::mutate(metric = "30-month source anomaly (z-score)")
rain_whiplash_df <- raster_mean_series(rain_whiplash_trough_to_peak) |> dplyr::mutate(metric = "Whiplash (z-score units)")

rain_whiplash_combined <- dplyr::bind_rows(rain_whiplash_source_df, rain_whiplash_df) |>
  dplyr::mutate(metric = factor(metric, levels = c("30-month source anomaly (z-score)", "Whiplash (z-score units)")))

ggplot(rain_whiplash_combined, aes(date, value)) +
  geom_hline(data = data.frame(metric = factor("30-month source anomaly (z-score)", levels = levels(rain_whiplash_combined$metric))),
             aes(yintercept = 0), colour = "grey70", linewidth = 0.3, inherit.aes = FALSE) +
  geom_line(colour = whiplash_colour) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = NULL)
```

```{r}
#| label: fig-rain-seasonal
#| fig-cap: "Rainfall's per-season total (a sum, not a rolling average) and its own anomaly, national mean, at each season-end month."
#| fig-asp: 0.55

rain_seasonal_df <- raster_mean_series(rain_seasonal_raster) |> dplyr::mutate(window = "Seasonal", metric = "Level (mm, per-season total)")
rain_seasonal_anom_df <- raster_mean_series(rain_seasonal_anomaly_raster) |> dplyr::mutate(window = "Seasonal", metric = "Anomaly (z-score)")

rain_seasonal_combined <- dplyr::bind_rows(rain_seasonal_df, rain_seasonal_anom_df) |>
  dplyr::mutate(metric = factor(metric, levels = c("Level (mm, per-season total)", "Anomaly (z-score)")))

ggplot(rain_seasonal_combined, aes(date, value)) +
  geom_hline(data = data.frame(metric = factor("Anomaly (z-score)", levels = levels(rain_seasonal_combined$metric))),
             aes(yintercept = 0), colour = "grey70", linewidth = 0.3, inherit.aes = FALSE) +
  geom_line(colour = window_colours[["Seasonal"]]) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = NULL)
```

### Where is this changing most? {#sec-rain-zones}

```{r}
#| label: fig-rain-zone-anomaly
#| fig-cap: "Rainfall anomaly (12-month rolling, z-score) by AE zone -- winter-dominant rainfall in the south vs. summer-dominant rainfall in the north means a national mean can mix two quite different stories together; this is the check for that. Shared y-axis across facets, since anomaly is already standardised."
#| fig-asp: 1.1

rain_zone_anomaly_df <- raster_zone_series(rain_rolling_anomaly_raster_12, aez_adj) |>
  dplyr::left_join(zone_regions, by = "ae_zone")

ggplot(rain_zone_anomaly_df, aes(date, value, colour = ae_zone)) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.3) +
  geom_line(linewidth = 0.35) +
  facet_wrap(~region_label, ncol = 2) +
  scale_colour_manual(values = zone_colours) +
  labs(x = NULL, y = "Rainfall anomaly (z-score)", colour = "AE zone") +
  theme(legend.position = "right", legend.text = element_text(size = 8), strip.text = element_text(size = 8.5))
```

```{r}
#| label: fig-rain-zone-whiplash
#| fig-cap: "Rainfall whiplash (trough-to-peak) by AE zone, same grouping as above."
#| fig-asp: 1.1

rain_zone_whiplash_df <- raster_zone_series(rain_whiplash_trough_to_peak, aez_adj) |>
  dplyr::left_join(zone_regions, by = "ae_zone")

ggplot(rain_zone_whiplash_df, aes(date, value, colour = ae_zone)) +
  geom_line(linewidth = 0.35) +
  facet_wrap(~region_label, ncol = 2) +
  scale_colour_manual(values = zone_colours) +
  labs(x = NULL, y = "Rainfall whiplash (z-score units)", colour = "AE zone") +
  theme(legend.position = "right", legend.text = element_text(size = 8), strip.text = element_text(size = 8.5))
```

### How redundant are the rainfall versions?

```{r}
#| label: rain-correlation

rain_cor <- paddock_season_covs |>
  dplyr::select(rain_rolling3, rain_rolling6, rain_rolling12, rain_seasonal,
                rain_whiplash_trough_to_peak) |>
  tidyr::drop_na() |>
  cor() |>
  round(2)

rain_cor |> as.data.frame() |> gt::gt(rownames_to_stub = TRUE) |> gt::tab_options(table.font.size = "0.8em")
```

## Soil moisture {#sec-soilmoisture}

**Source:** AWRA-L v7's rootzone soil moisture (`sm_pct`, BOM/CSIRO's Australian Water Resources Assessment model, 1911+), served as one whole-history file, updated periodically as new months are appended.
**Native resolution:** ~5 km (resampled onto PML's coarser shared 0.1&deg; grid before any temporal processing).
**Temporal span:** 1982+ (trimmed from AWRA's own full 1911+ history to match the other covariates' common start).

**Processing:** resample onto PML's 0.1&deg; grid &rarr; trim to 1982+ &rarr; used directly, with **no rolling window or seasonal summing** -- AWRA's own water-balance model has already integrated rainfall/evapotranspiration/drainage history into this single value, so any further windowing would double-count memory already built in. Leave-one-out z-score anomaly, purely per-pixel (no neighbour-blending), plus a **whiplash** version built the same way as rainfall's own (its own dedicated 30-month rolling anomaly, @sec-whiplash) -- despite the double-counting concern just above, which this reverses; tested directly (@sec-whiplash-validation) rather than left as a theoretical worry, and the rolling average turned out to matter: a matching sweep with no rolling average at all scored meaningfully worse.

```{r}
#| label: fig-soilmoisture-timeseries
#| fig-cap: "Soil moisture: national mean of the raw value and its anomaly, at each season-end month."

sm_df <- raster_mean_series(soil_moisture_raster_coarse) |> dplyr::rename(soil_moisture = value)
sm_anom_df <- raster_mean_series(soil_moisture_anomaly_raster) |> dplyr::rename(soil_moisture_anomaly = value)

sm_combined <- dplyr::full_join(sm_df, sm_anom_df, by = "date") |>
  tidyr::pivot_longer(-date, names_to = "series", values_to = "value")

ggplot(sm_combined, aes(date, value)) +
  geom_line(colour = "#2166ac") +
  facet_wrap(~series, scales = "free_y", ncol = 1,
             labeller = as_labeller(c(soil_moisture = "Soil moisture (% full)", soil_moisture_anomaly = "Soil moisture anomaly (z-score)"))) +
  labs(x = NULL, y = NULL)
```

```{r}
#| label: fig-soilmoisture-whiplash
#| fig-cap: "Soil moisture's whiplash source (its own dedicated 30-month rolling anomaly, top) and the resulting whiplash swing (bottom), national mean -- same layout as rainfall's own whiplash figure above."
#| fig-asp: 0.6

sm_whiplash_source_df <- raster_mean_series(soil_moisture_whiplash_source_anomaly) |> dplyr::mutate(metric = "30-month source anomaly (z-score)")
sm_whiplash_df <- raster_mean_series(soil_moisture_whiplash_trough_to_peak) |> dplyr::mutate(metric = "Whiplash (z-score units)")

sm_whiplash_combined <- dplyr::bind_rows(sm_whiplash_source_df, sm_whiplash_df) |>
  dplyr::mutate(metric = factor(metric, levels = c("30-month source anomaly (z-score)", "Whiplash (z-score units)")))

ggplot(sm_whiplash_combined, aes(date, value)) +
  geom_hline(data = data.frame(metric = factor("30-month source anomaly (z-score)", levels = levels(sm_whiplash_combined$metric))),
             aes(yintercept = 0), colour = "grey70", linewidth = 0.3, inherit.aes = FALSE) +
  geom_line(colour = whiplash_colour) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = NULL)
```

```{r}
#| label: fig-soilmoisture-zone-anomaly
#| fig-cap: "Soil moisture anomaly (z-score) by AE zone, same grouping as GPP/rainfall's own zone figures above. Shared y-axis across facets, since anomaly is already standardised."
#| fig-asp: 1.1

sm_zone_anomaly_df <- raster_zone_series(soil_moisture_anomaly_raster, aez_adj) |>
  dplyr::left_join(zone_regions, by = "ae_zone")

ggplot(sm_zone_anomaly_df, aes(date, value, colour = ae_zone)) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.3) +
  geom_line(linewidth = 0.35) +
  facet_wrap(~region_label, ncol = 2) +
  scale_colour_manual(values = zone_colours) +
  labs(x = NULL, y = "Soil moisture anomaly (z-score)", colour = "AE zone") +
  theme(legend.position = "right", legend.text = element_text(size = 8), strip.text = element_text(size = 8.5))
```

```{r}
#| label: fig-soilmoisture-zone-whiplash
#| fig-cap: "Soil moisture whiplash (trough-to-peak) by AE zone, same grouping as above."
#| fig-asp: 1.1

sm_zone_whiplash_df <- raster_zone_series(soil_moisture_whiplash_trough_to_peak, aez_adj) |>
  dplyr::left_join(zone_regions, by = "ae_zone")

ggplot(sm_zone_whiplash_df, aes(date, value, colour = ae_zone)) +
  geom_line(linewidth = 0.35) +
  facet_wrap(~region_label, ncol = 2) +
  scale_colour_manual(values = zone_colours) +
  labs(x = NULL, y = "Soil moisture whiplash (z-score units)", colour = "AE zone") +
  theme(legend.position = "right", legend.text = element_text(size = 8), strip.text = element_text(size = 8.5))
```

## Minimum temperature {#sec-mintemp}

**Source:** SILO's daily minimum temperature (Bureau of Meteorology, 1980+) -- no pre-aggregated monthly/seasonal product exists for this variable, unlike rainfall, so daily files are downloaded and aggregated by this pipeline itself, one netCDF per calendar year (`download_silo_daily_data()`).
**Native resolution:** 0.05&deg; -- kept at SILO's own native grid, not resampled onto PML's 0.1&deg; grid like the other three: min_temp is extracted at each paddock's own point rather than compared pixel-for-pixel against them (see [Attaching to the seasonal model](#sec-attach) below).
**Temporal span:** 1980-`r max(unlist(min_temp_year_blocks))` (5-year blocks).

**Processing:** aggregate daily composites to monthly means in `r length(min_temp_year_blocks)` independent, fixed 5-year blocks spanning `r min(unlist(min_temp_year_blocks))`-`r max(unlist(min_temp_year_blocks))` (`build_year_blocks()`, `summarise_silo_to_month()`), not as one command covering the whole record &rarr; concatenate every block into one continuous stack (`combine_spatraster_list()`) &rarr; average within each of the 4 seasons (`build_seasonal_raster()`). **Deliberately kept as a raw value, never converted to an anomaly** -- winter cold is used directly as a breeding-relevant threshold, not as a deviation from that paddock's own long-term norm.

::: {.callout-note}
#### Why blocks, not one 46-year command

SILO's daily NetCDF files store each day as one indivisible chunk covering the *entire country* -- there's no way to decompress a spatial subset without loading the whole day, and processing all `r max(unlist(min_temp_year_blocks)) - min(unlist(min_temp_year_blocks)) + 1` years at once as a single `terra::tapp()` call repeatedly exhausted memory and was killed partway through, hours in. Splitting the record into fixed 5-year blocks bounds peak memory to one block at a time, lets independent blocks build concurrently, and means a killed block only costs re-running that one block -- not the whole record.
:::

```{r}
#| label: fig-mintemp-blocks
#| fig-cap: "Minimum temperature's daily SILO record is aggregated in independent 5-year blocks (not one command spanning the whole record) -- each block builds and caches on its own."
#| fig-asp: 0.22

mt_blocks_df <- purrr::imap_dfr(min_temp_year_blocks, ~ tibble::tibble(block = .y, start = min(.x), end = max(.x)))

ggplot(mt_blocks_df, aes(xmin = start, xmax = end + 1, ymin = 0, ymax = 1, fill = block)) +
  geom_rect(colour = "white", show.legend = FALSE) +
  geom_text(aes(x = (start + end + 1) / 2, y = 0.5, label = paste0(start, "-", end)), size = 3, colour = "white") +
  scale_fill_brewer(palette = "Paired") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10), name = NULL) +
  labs(y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

```{r}
#| label: fig-mintemp-timeseries
#| fig-cap: "Minimum temperature: national mean seasonal average, all 4 seasons (not just winter)."

## raster_mean_series() assigns each season a fixed representative month
## (Summer -> Jan, Autumn -> Apr, Winter -> Jul, Spring -> Oct) -- reverse
## that same mapping here rather than re-deriving season from the raster's
## own (differently-ordered) layer names, which would risk misaligning once
## the series has been sorted by date.
mt_df <- raster_mean_series(min_temp_seasonal_raster) |>
  dplyr::mutate(
    season = factor(
      c(`1` = "Summer", `4` = "Autumn", `7` = "Winter", `10` = "Spring")[as.character(lubridate::month(date))],
      levels = c("Summer", "Autumn", "Winter", "Spring")
    )
  )

ggplot(mt_df, aes(date, value, colour = season)) +
  geom_line(aes(group = 1), colour = "grey70") +
  geom_point(size = 1.5) +
  labs(x = NULL, y = "Mean minimum temperature (°C)", colour = "Season") +
  scale_colour_manual(values = c(Summer = "#e41a1c", Autumn = "#ff7f00", Winter = "#377eb8", Spring = "#4daf4a"))
```

```{r}
#| label: fig-mintemp-zone
#| fig-cap: "Minimum temperature, winter only (the breeding-relevant season this covariate is actually meant to capture), by AE zone -- the national mean above collapses away exactly the latitudinal spread that matters here. Shared y-axis across facets: absolute degrees are directly comparable zone to zone, unlike a standardised anomaly."
#| fig-asp: 1.1

mt_zone_df <- raster_zone_series(min_temp_seasonal_raster, aez_adj) |>
  dplyr::filter(lubridate::month(date) == 7) |> # winter's own representative month (see raster_mean_series())
  dplyr::left_join(zone_regions, by = "ae_zone")

ggplot(mt_zone_df, aes(date, value, colour = ae_zone)) +
  geom_line(linewidth = 0.35) +
  facet_wrap(~region_label, ncol = 2) +
  scale_colour_manual(values = zone_colours) +
  labs(x = NULL, y = "Winter mean minimum temperature (°C)", colour = "AE zone") +
  theme(legend.position = "right", legend.text = element_text(size = 8), strip.text = element_text(size = 8.5))
```

```{r}
#| label: fig-mintemp-map
#| fig-cap: "Minimum temperature, most recent instance of each season: the latitudinal breeding-threshold gradient this covariate is meant to capture is a spatial pattern by nature, so unlike every other covariate above, the most informative view of min_temp is a map even at a single point in time, not a national mean series. The zone lines above show each zone's own winter trend over time; this map shows the full spatial gradient at the current moment -- complementary, not redundant."
#| fig-asp: 0.65

mt_labels <- names(min_temp_seasonal_raster)
mt_seasons_all <- sub("-.*", "", mt_labels)
mt_years_all   <- as.integer(sub(".*-", "", mt_labels))
mt_latest_layers <- sapply(c("Summer", "Autumn", "Winter", "Spring"), function(s) {
  idx <- which(mt_seasons_all == s)
  mt_labels[idx[which.max(mt_years_all[idx])]]
})

mt_map_df <- raster_map_df(min_temp_seasonal_raster[[mt_latest_layers]])

ggplot(mt_map_df, aes(x, y, fill = value)) +
  geom_raster() +
  facet_wrap(~label) +
  coord_equal() +
  scale_fill_viridis_c(option = "B", name = "Min. temp (°C)") +
  labs(x = NULL, y = NULL) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

## Attaching to the seasonal model {#sec-attach}

Every covariate above is ultimately read onto `paddock_season_grid` (one row per paddock x season) via one of two matching strategies, depending on how often each covariate actually produces a new value:

- **GPP's rolling-3/rolling-6/rolling-12 rasters, rainfall's rolling-3/rolling-6/rolling-12, soil moisture, and both whiplash rasters (rainfall/soil moisture)** are still built at monthly- or season-end grain internally, but only *read* once per season, at that season's own last calendar month (`period_col = "season_end_month_year"`, e.g. Winter reads August's value) -- not averaged further across the season, since all of these are already a point-in-time value. This works unchanged for whiplash's monthly-grain source (2026-08): a monthly raster's own layer names (`"<month>_<year>"`) already match `season_end_month_year`'s own format directly, so no special-casing was needed to attach it the same way as everything else.
- **Rainfall's per-season total and min temperature** already produce a real, distinct value for every season directly (`period_col = "season_year_adj"`), so no broadcasting is needed. GPP has no such version any more -- its shortest window (3-month) is already read at `"season_end_month_year"` grain alongside its longer windows, above.

```{r}
#| label: fig-completeness
#| fig-cap: "Data completeness (% non-missing) by covariate and season type -- gaps reflect each covariate's own warm-up period (e.g. GPP's 12-month rolling window needs 12 months of history before its first value; whiplash needs its own 30-month source's warm-up plus a further 12-month sweep on top of that; leave-one-out anomalies need at least 3 years of record)."
#| fig-asp: 1.2

completeness <- paddock_season_covs |>
  dplyr::select(
    season,
    gpp_rolling3, gpp_rolling3_anomaly, gpp_rolling6, gpp_rolling6_anomaly, gpp, gpp_anomaly,
    rain_rolling3, rain_rolling3_anomaly, rain_rolling6, rain_rolling6_anomaly, rain_rolling12, rain_rolling12_anomaly,
    rain_seasonal, rain_seasonal_anomaly, rain_whiplash_trough_to_peak,
    soil_moisture, soil_moisture_anomaly, soil_moisture_whiplash_trough_to_peak,
    min_temp
  ) |>
  dplyr::group_by(season) |>
  dplyr::summarise(dplyr::across(everything(), ~ mean(!is.na(.x)) * 100), .groups = "drop") |>
  tidyr::pivot_longer(-season, names_to = "covariate", values_to = "pct_complete")

ggplot(completeness, aes(x = covariate, y = pct_complete, fill = season)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = "% complete", fill = "Season") +
  scale_fill_manual(values = c(Summer = "#e41a1c", Autumn = "#ff7f00", Winter = "#377eb8", Spring = "#4daf4a"))
```

## Reproducing this from scratch {#sec-reproduce}

Every figure and table above is built directly from `_targets.R` (section B) targets, `tar_load()`'d at the top of this document (see the `setup` chunk) -- rerun `targets::tar_make()` after any raw-data or code change, then re-render this document, to regenerate everything with the refreshed values. No numbers here are hand-entered; the summary table, bias-correction table and "as of" dates above all read straight off the live pipeline output.

```{r}
#| label: pipeline-reference

pipeline_ref <- tibble::tribble(
  ~Covariate,        ~`Key _targets.R targets (in build order)`,                                                                                              ~`Key r/ functions`,
  "GPP",             "A) NASA (self-contained): gpp_files_modis, gpp_files_viirs_frozen_block, gpp_files_viirs_recent_block, gpp_files_viirs_nongf -> composite_gpp_rast_modis/_viirs_gf/_viirs_nongf (composite grain) -> gpp_modis_pixel_ratio_rast, gpp_gf_pixel_ratio_rast -> composite_gpp_rast_modis_corrected, composite_gpp_rast_viirs_nongf_corrected -> composite_gpp_rast (spliced) -> monthly_gpp_rast. B) PML (self-contained): pml_gpp_files -> pml_gpp_rast. C) Combine: monthly_gpp_rast, pml_gpp_rast -> monthly_gpp_rast_coarse -> gpp_pml_pixel_ratio_rast -> pml_gpp_rast_corrected -> gpp_raster_coarse -> {tar_map() over gpp_rolling_windows: gpp_rolling_raster_3/_6/_12 -> gpp_rolling_anomaly_raster_3/_6/_12}. No whiplash (tried and dropped 2026-08 -- no significant real-plague onset-forecasting result at any of 126 tested source/sweep/anchor combinations, see @sec-whiplash-validation). Splice/correction methodology (gpp_pml_bias_check, gpp_gf_bias_check, gpp_modis_bias_check, gpp_gf_pixel_ratio_rast, gpp_modis_pixel_ratio_rast) covered in gpp_processing.qmd, not here.", "download_gpp(), download_gpp_block(), load_and_clamp_gpp_composites(), finish_gpp_period_raster(), compare_gpp_rasters(), resample_to_grid(), build_gpp_pixel_ratio_raster(), apply_gpp_pixel_ratio_correction(), combine_spatraster_list(), compute_rolling_mean_raster(), compute_loo_anomaly_raster(), trim_to_season_end_months()",
  "Rainfall",        "silo_files -> rainfall_raster -> rain_raster_coarse -> {tar_map() over rain_rolling_windows: rain_rolling_raster_3/_6/_12 -> rain_rolling_anomaly_raster_3/_6/_12} (level/anomaly comparison only) + rain_seasonal_raster -> rain_seasonal_anomaly_raster; separately, rain_raster_coarse -> rain_whiplash_source_raster (30-month, monthly grain) -> rain_whiplash_source_anomaly -> rain_whiplash_trough_to_peak (12-month sweep, anchor='current', 2026-08, validated against real plague occurrence -- @sec-whiplash-validation)", "download_silo_monthly_data(), build_monthly_rainfall_raster(), resample_to_grid(), compute_rolling_mean_raster(), build_seasonal_raster(), compute_loo_anomaly_raster(), compute_whiplash_raster(), trim_to_season_end_months()",
  "Soil moisture",   "awra_files -> soil_moisture_raster -> soil_moisture_raster_coarse -> soil_moisture_anomaly_raster; separately, soil_moisture_raster -> soil_moisture_whiplash_source_raster (30-month, monthly grain, its own resample+trim since soil_moisture_raster_coarse is already season-trimmed) -> soil_moisture_whiplash_source_anomaly -> soil_moisture_whiplash_trough_to_peak (12-month sweep, anchor='current', 2026-08, validated against real plague occurrence -- @sec-whiplash-validation)", "download_awra_data(), build_awra_soil_moisture_raster(), resample_to_grid(), compute_loo_anomaly_raster(), compute_rolling_mean_raster(), compute_whiplash_raster(), trim_to_season_end_months()",
  "Min temperature", "min_temp_year_blocks -> silo_daily_files_min_temp -> min_temp_raster_block -> min_temp_raster -> min_temp_raster_coarse -> min_temp_seasonal_raster",                "download_silo_daily_data(), build_year_blocks(), summarise_silo_to_month(), combine_spatraster_list(), resample_to_grid(), build_seasonal_raster()",
  "Attach (all)",    "paddock_season_grid -> paddock_season_covs",                                                                                              "build_paddock_season_grid(), attach_raster_covs()"
)

pipeline_ref |>
  gt::gt() |>
  gt::tab_options(table.font.size = "0.72em") |>
  gt::cols_width(Covariate ~ px(110))
```
