---
title: "House mouse forecast"
subtitle: Winter 2025
description: "Regional seasonal forecast of house mouse abundance in the Australian grainbelt"
#date: last-modified
title-block-banner: "white"
#title-block-banner-color: "black" # title font colour
format: 
  html:
    self-contained: true
    theme: cosmo
    fontcolor: black
    fig-width: 8  
    fig-asp: 0.8 # fig-height = fig-width * fig-asp
css: styles.css
execute:
  echo: false
  warning: false
  message: false
editor: source
toc: TRUE
---

```{r}
#| label: setup
#| message: false
#| include: FALSE

## load packages
library(targets)
library(dplyr)
library(sf)
library(ggplot2)
library(flextable)
library(scico)
library(viridis)

# set the default theme for ggplot objects
theme_update(panel.grid = element_blank(),
            legend.position = "bottom")
```

```{r}
#| label: load
#| message: false
#| include: FALSE

## load targets
targets::tar_load(
  c(model_data,
    model_predictions_ae_zone,
    model_predictions_region,
    model_predictions_trend,
    recent_survey_stats_preds))

# read in australian outline shapefile and adjust depending on how data has been filtered 
aus_shp = sf::read_sf("raw_data/predictor_variables/australian_borders/aus_outline_states.shp") |> 
    sf::st_transform(crs = "EPSG:4326") |> 
    dplyr::filter(ST_NAME != "WA")
  
# town / city centres for labelling
town_labels = sf::st_as_sf(read.csv("raw_data/town_labels.csv"), coords = c("longitude", "latitude"), crs = "EPSG:4326", remove = FALSE)
```

```{r}
#| label: split_season_summaries
#| message: false
#| include: FALSE

# recent_survey_stats_preds is a list with $ae_zone and $region level summaries
# use ae_zone level for the "In short" overview, region level for detailed tables
df_aez <- recent_survey_stats_preds$ae_zone
df_region <- recent_survey_stats_preds$region

# split into current and next season using season_year_adj ordering
all_seasons <- sort(unique(df_aez$season_year_adj))
current_season <- all_seasons[length(all_seasons) - 1]
next_season <- all_seasons[length(all_seasons)]

df_current <- filter(df_aez, season_year_adj == current_season)
df_next <- filter(df_aez, season_year_adj == next_season)

```


```{r}
#| label: site_data

site_data_for_plot <- model_data |>

  # remove unsurveyed seasons for each site
  filter(!(is.na(mice_prop))) |>

  # take relevant columns
  dplyr::select(longitude, latitude, season, season_year_adj, mice_prop) |>

  # new column for whether any mice signs were detected
  group_by(longitude, latitude, season_year_adj) %>%
  mutate(mice01 = if_else(sum(mice_prop) > (0.001 * n()), "yes", "no")) |>
  ungroup() |>

  # just one row per site, subsite, season_year_adj
  unique() %>%

  # make spatial for plotting
  sf::st_as_sf(., coords = c("longitude", "latitude"), crs = "EPSG:4326") %>%

  # align season_year_adj factor levels with prediction data
  dplyr::mutate(season_year_adj = factor(season_year_adj, levels = levels(model_predictions_ae_zone$season_year_adj)))

```



```{r}
#| label: prediction_data

# filter prediction data to the most recent 2 seasons (current + forecast) at each level
prediction_data_aez <- model_predictions_ae_zone |>
   filter(time %in% (max(time)-1):(max(time)))

prediction_data_region <- model_predictions_region |>
   filter(time %in% (max(time)-1):(max(time)))

prediction_data_trend <- model_predictions_trend |>
   filter(time %in% (max(time)-1):(max(time)))

# helper: build a map with layers in correct order:
#   1. Australian outline (background)  2. prediction data  3. town labels  4. map crop
# data_layer should be a geom_sf call + scale, which gets sandwiched between background and labels
build_map <- function(data_layer) {
  ggplot() +
    # background
    geom_sf(data = aus_shp, fill = "white") +
    labs(y = "", x = "") +
    # prediction polygons (added in the middle so they sit on top of outline but below labels)
    data_layer +
    # town labels on top
    geom_sf(data = town_labels, size = 0.75) +
    ggrepel::geom_text_repel(
      data = town_labels, aes(longitude, latitude, label = name), size = 2.5,
      #Ceduna Jamestown Yorketown Adelaide Horsham Birchip Melbourne Bendigo Albury Wagga Sydney Griffith Dubbo Walgett Gunnedah Goondiwindi Brisbane Dalby  St George
      nudge_x = c(0.0, 0.0, -3.2, -0.5, -2.0, -1.5, -1.0, 1.0, 1.0, 0.0, 1.5, -1.0, -3.0, -3.0, 0.0, -8.0, 0.0, -2.0, -5.0),
      nudge_y = c(1.5, 2.0, -1.0, -2.0, -4.0, 1.0, -2.0, -1.0, -1.5, 1.5, -1.1, -1.0, 1.0, 0.0, 1.5, 0.0, -1.5, 3.0, 1.5)) +
    coord_sf(xlim = c(128, 155), ylim = c(-40, -24), expand = FALSE)
}
```

## In short

+ `r {
  levels_present <- unique(df_current$cat3[df_current$cat3 %in% c("Low", "Moderate", "High")])
  ordered_levels <- c("Low", "Moderate", "High")
  present_ordered <- ordered_levels[ordered_levels %in% levels_present]
  
  if (length(present_ordered) == 1) {
    paste0("The model predicts current mouse abundance to be ", tolower(present_ordered), ".")
  } else if (length(present_ordered) == 2) {
    paste0("The model predicts current mouse abundance to range from ", 
           tolower(present_ordered[1]), " to ", tolower(present_ordered[2]), ".")
  } else if (length(present_ordered) == 3) {
    "The model predicts current mouse abundance to range from low to high."
  } else {
    "Mouse abundance predictions are not available."
  }
}`


+ `r {
  moderate_zones <- df_current$ae_zone[df_current$cat3 == "Moderate"]
  high_zones <- df_current$ae_zone[df_current$cat3 == "High"]

  list_zones <- function(zones) {
    if (length(zones) == 0) return(NULL)
    if (length(zones) == 1) return(zones)
    paste(paste(zones[-length(zones)], collapse = ", "), "and", zones[length(zones)])
  }

  parts <- c()
  if (length(moderate_zones) > 0) {
    parts <- c(parts, paste0("moderate in ", list_zones(moderate_zones)))
  }
  if (length(high_zones) > 0) {
    parts <- c(parts, paste0("high in ", list_zones(high_zones)))
  }

  if (length(parts) > 0) {
    paste0("Mouse abundance is predicted to currently be ", paste(parts, collapse = " and "), ".")
  } else {
    "No agroecological zones are predicted to currently have moderate or high mouse abundance."
  }
}`


+ `r {
  increase_df <- df_next[df_next$Prop_Change > 1.25, c("ae_zone", "Prop_Change")]

  if (nrow(increase_df) == 0) {
    "Mouse abundance is not forecasted to increase in any zones next season."
  } else {
    # Format with integer % (no decimals)
    increase_df$label <- paste0(
      increase_df$ae_zone,
      " (by ",
      round((increase_df$Prop_Change - 1) * 100, 0),  # Round to whole number
      "%)"
    )

    # Join with commas and 'and'
    if (nrow(increase_df) == 1) {
      sentence <- increase_df$label
    } else {
      sentence <- paste(
        paste(increase_df$label[-nrow(increase_df)], collapse = ", "),
        "and",
        increase_df$label[nrow(increase_df)]
      )
    }

    paste0(
      "Mouse abundance is forecasted to increase next season in ",
      sentence, "."
    )
  }
}`
`r {
  decrease_df <- df_next[df_next$Prop_Change < 0.75, c("ae_zone", "Prop_Change")]

  if (nrow(decrease_df) == 0) {
    "Mouse abundance is not forecasted to decrease in any zones next season."
  } else {
    # Format zone names with integer % decrease
    decrease_df$label <- paste0(
      decrease_df$ae_zone,
      " (by ",
      abs(round((1 - decrease_df$Prop_Change) * 100, 0)),
      "%)"
    )

    # Join zones with commas and "and"
    if (nrow(decrease_df) == 1) {
      sentence <- decrease_df$label
    } else {
      sentence <- paste(
        paste(decrease_df$label[-nrow(decrease_df)], collapse = ", "),
        "and",
        decrease_df$label[nrow(decrease_df)]
      )
    }

    paste0(
      "Mouse abundance is forecasted to decrease next season in ",
      sentence, "."
    )
  }
}`


+ `r {
  min_prob <- min(df_next$pr_threshold, na.rm = TRUE)
  max_prob <- max(df_next$pr_threshold, na.rm = TRUE)
  paste0("The probability of a mouse plague next season ranges from ",
         round(min_prob * 100), "% to ", round(max_prob * 100), "% across agroecological zones.")
}`


```{r}
#| label: plot_3cat_helper
#| include: false

# helper: 3-category abundance map (Low / Moderate / High)
plot_3cat <- function(data, alpha = 1) {
  build_map(list(
    geom_sf(data = data,
            aes(fill = factor(cat3, levels = c("Low", "Moderate", "High"), ordered = TRUE)),
            alpha = alpha, show.legend = TRUE),
    scale_fill_manual(values = c("Low" = "#91cf60", "Moderate" = "#ffeda0", "High" = "#f03b20"),
                      breaks = c("Low", "Moderate", "High"),
                      name = "Relative abundance",
                      drop = FALSE)
  ))
}
```

::: {.panel-tabset}

#### AE zone

```{r}
#| label: fig_3cat_aez
#| fig-cap:
#|   - "Predicted relative abundance of house mice in the current season, across modified GRDC agroecological zones."

plot_3cat(filter(prediction_data_aez, time == max(time) - 1))
```

#### Region

```{r}
#| label: fig_3cat_region
#| fig-cap:
#|   - "Predicted relative abundance of house mice in the current season, by region."

plot_3cat(filter(prediction_data_region, time == max(time) - 1))
```

#### Site

```{r}
#| label: fig_3cat_trend
#| fig-cap:
#|   - "Predicted relative abundance of house mice in the current season, by site (shared latent trend)."

plot_3cat(filter(prediction_data_trend, time == max(time) - 1), alpha = 0.7)
```

:::


### Management reccomendations

Mouse numbers are moderate to high in many areas (highly patchy). Given the excellent conditions across many
areas of southern Queensland, northern NSW and parts of WA, and in parts of Eyre Peninsula (SA; which was
affected by hail), mice have been breeding through summer and autumn, which will peak in March/April at the time
of sowing winter crops. See GRDC Mouse Control website for more details about control options. Be aware that
only ZnP25 baits are currently available (Emergency Permits previously issued for ZnP50 have expired).

1. Actively gauge mouse numbers by walking through paddocks (mouse chew cards and active burrow counts are
useful tools at this time of year).
2. At 6 weeks prior to sowing (now): If mouse activity is high and there is plenty of background food, reduce the
background food through light tillage or sheep grazing, and strategically bait with ZnP25 as the only currently
registered ZnP bait available. If mice are still a problem at sowing, be prepared to bait again at that time,
ensuring a 6-week gap to overcome any behavioural aversion from ingesting a sub-lethal dose.
3. At sowing: If mice remain a problem at sowing, apply bait off the back of the seeder to prevent damage to the
freshly sown crop.
4. Baiting at sowing is most effective if no other food sources are available.
5. Talk to bait suppliers early to ensure timely supply. As with any agricultural chemical, use the product in
accordance with the label instructions, and report any adverse or off target effects via the APVMA website.


```{r}
#| label: plot_helpers
#| include: false

# shared viridis colour scale limits across all spatial levels (for comparable maps)
pred_range <- range(log(c(model_predictions_ae_zone$pred_mean,
                          model_predictions_region$pred_mean,
                          model_predictions_trend$pred_mean)), na.rm = TRUE)

# helper: continuous abundance map
plot_abundance <- function(data, alpha = 1) {
  build_map(list(
    geom_sf(data = data, aes(fill = log(pred_mean)), alpha = alpha),
    scale_fill_viridis_c(
      limits = pred_range,
      breaks = pred_range,
      labels = c("Low", "High"),
      name = "Relative abundance\n(log-scaled)")
  ))
}

# helper: proportional change map
plot_change <- function(data, alpha = 1) {
  build_map(list(
    geom_sf(data = data, aes(fill = Prop_Change), alpha = alpha),
    scale_fill_scico(palette = "vik", midpoint = 1, name = "Proportional change\nfrom current season")
  ))
}

# helper: outbreak probability map
plot_probability <- function(data, alpha = 1) {
  build_map(list(
    geom_sf(data = data, aes(fill = pr_threshold * 100), alpha = alpha),
    scale_fill_scico(
      palette = "vik",
      limits = c(0, 100),
      breaks = c(0, 25, 50, 75, 100),
      name = "Outbreak probability (%)")
  ))
}
```


## In detail

### Predicted current situation

::: {.panel-tabset}

#### AE zone

```{r}
#| label: plot_current_aez
#| fig-cap:
#|   - "Predicted relative abundance of house mice in the current season, across modified GRDC agroecological zones."

plot_abundance(filter(prediction_data_aez, time == max(time) - 1))
```

#### Region

```{r}
#| label: plot_current_region
#| fig-cap:
#|   - "Predicted relative abundance of house mice in the current season, by region."

plot_abundance(filter(prediction_data_region, time == max(time) - 1))
```

#### Site

```{r}
#| label: plot_current_trend
#| fig-cap:
#|   - "Predicted relative abundance of house mice in the current season, by site (shared latent trend)."

plot_abundance(filter(prediction_data_trend, time == max(time) - 1), alpha = 0.7)
```

:::


### Forecasted change next season

::: {.panel-tabset}

#### AE zone

```{r}
#| label: plot_change_aez
#| fig-cap:
#|   - "Predicted change in relative abundance of house mice from the current to the future season, across modified GRDC agroecological zones."

plot_change(filter(prediction_data_aez, time == max(time)))
```

#### Region

```{r}
#| label: plot_change_region
#| fig-cap:
#|   - "Predicted change in relative abundance by region."

plot_change(filter(prediction_data_region, time == max(time)))
```

#### Site

```{r}
#| label: plot_change_trend
#| fig-cap:
#|   - "Predicted change in relative abundance by site (shared latent trend)."

plot_change(filter(prediction_data_trend, time == max(time)), alpha = 0.7)
```

:::


### Probability of plague event

::: {.panel-tabset}

#### AE zone

```{r}
#| label: plot_prob_aez
#| fig-cap:
#|   - "Predicted probability of mouse abundance being in a high state, across modified GRDC agroecological zones."

plot_probability(prediction_data_aez) + facet_wrap(~season)
```

#### Region

```{r}
#| label: plot_prob_region
#| fig-cap:
#|   - "Predicted probability of mouse abundance being in a high state, by region."

plot_probability(prediction_data_region) + facet_wrap(~season)
```

#### Site

```{r}
#| label: plot_prob_trend
#| fig-cap:
#|   - "Predicted probability of mouse abundance being in a high state, by site (shared latent trend)."

plot_probability(prediction_data_trend, alpha = 0.7) + facet_wrap(~season)
```

:::



### Data summary

::: {.panel-tabset}

#### AE zone

```{r}
#| label: data_tbl_aez
#| tbl-cap:
#|   - "Summary table of raw data and model predictions per agroecological zone."

# prepare ae_zone level data for display, newest season first then by ae_zone
tbl_aez <- df_aez |>
  dplyr::arrange(desc(season_year_adj), ae_zone) |>
  dplyr::select(-any_of(c("season_year_adj", "pred_median")))

  # change defaults
  set_flextable_defaults(
    digits = 2,
    font.size = 10,
    na_str = "NA",
    theme_fun = theme_vanilla,
    padding = 6,
    background.color = "#EFEFEF")


  ft_aez <- flextable(tbl_aez) |>

    # merge cells for year, season, and ae_zone so not repeated
    merge_v(j = c("year_adj", "season", "ae_zone")) |>

    # centre all values
    align(align = "center", part = "all") |>

    # limit to 2 decimal points for these numeric columns
    colformat_double(j = c("prop_traps", "pred_mean", "pred_lower", "pred_upper", "Prop_Change", "pr_threshold"), digits = 2) |>

    # rename columns
    set_header_labels(
      year_adj = "Year",
      season = "Season",
      ae_zone = "AE zone",
      sites_surveyed_burrows = "Sites",
      sites_1burrow = "Active sites",
      sites_surveyed_chew = "Sites",
      sites_1chew = "Active sites",
      sites_surveyed_traps = "Sites",
      prop_traps = "Proportion captures",
      pred_mean = "Mean",
      pred_lower = "Lower 95% CI",
      pred_upper = "Upper 95% CI",
      Prop_Change = "Change",
      pr_threshold = "Pr(plague)",
      cat3 = "Abundance") |>

    # add another header above column names (colwidths must sum to total number of columns)
    add_header_row(values = c("", "Burrows", "Chewcards", "Live-traps", "Model Predictions"),
                          colwidths = c(3, 2, 2, 2, 6), top = TRUE)

  # display table
  autofit(ft_aez)

```

#### Region

```{r}
#| label: data_tbl_region
#| tbl-cap:
#|   - "Summary table of raw data and model predictions per region within each agroecological zone."

# prepare region level data for display, newest season first then by ae_zone / region
tbl_region <- df_region |>
  dplyr::arrange(desc(season_year_adj), ae_zone, region) |>
  dplyr::select(-any_of(c("season_year_adj", "pred_median")))


  ft_region <- flextable(tbl_region) |>

    # merge cells for year, season, ae_zone, region so not repeated
    merge_v(j = c("year_adj", "season", "ae_zone", "region")) |>

    # centre all values
    align(align = "center", part = "all") |>

    # limit to 2 decimal points for these numeric columns
    colformat_double(j = c("prop_traps", "pred_mean", "pred_lower", "pred_upper", "Prop_Change", "pr_threshold"), digits = 2) |>

    # rename columns
    set_header_labels(
      year_adj = "Year",
      season = "Season",
      ae_zone = "AE zone",
      region = "Region",
      sites_surveyed_burrows = "Sites",
      sites_1burrow = "Active sites",
      sites_surveyed_chew = "Sites",
      sites_1chew = "Active sites",
      sites_surveyed_traps = "Sites",
      prop_traps = "Proportion captures",
      pred_mean = "Mean",
      pred_lower = "Lower 95% CI",
      pred_upper = "Upper 95% CI",
      Prop_Change = "Change",
      pr_threshold = "Pr(plague)",
      cat3 = "Abundance") |>

    # add another header above column names (colwidths must sum to total number of columns)
    add_header_row(values = c("", "Burrows", "Chewcards", "Live-traps", "Model Predictions"),
                          colwidths = c(4, 2, 2, 2, 6), top = TRUE)

  # display table
  autofit(ft_region)

```

:::

