---
title: "CSIRO Mouse Survelliance"
subtitle: "2025 report"
date: last-modified
format:
  html:
    self-contained: true
    theme: cosmo
    fontcolor: black
    number-sections: false
    toc: false
crossref:
  fig-title: "Figure"
  tbl-title: "Table"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
execute:
  echo: false
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup
#| message: false
#| include: FALSE

## load packages - this doc needs to be standalone from targets pipeline
library(targets)
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(ggh4x)
library(leaflet)
library(sf)
library(flextable)
library(gt)
library(patchwork)

## load targets
targets::tar_load(c(
  data,
  aus_shp
  ))

## load compare_monitoring_schedule function
source("r/compare_monitoring_schedule.R")

## create monitoring comparison table
monitoring_comparison <- compare_monitoring_schedule(
  data_list = data,
  start_season_year = "summer-2025",
  end_season_year = "spring-2025"
)

  # change flextable defaults
  set_flextable_defaults(
    digits = 2,
    font.size = 10,
    na_str = "NA",
    theme_fun = theme_vanilla,
    padding = 6,
    background.color = "#EFEFEF")
  

```


```{r}

#| label: data_manip

# filter data to current year
data_year <- purrr::map(data, dplyr::filter, year_adj == 2025)

# make dataframes spatial
data_year_sf <- purrr::map(data_year, st_as_sf,  coords = c("longitude", "latitude"), crs = 4326)

# Create tbl_traps dataframe 
tbl_traps <- data_year$traps |>
  # arrange by time first
  arrange(season_year_adj, state, region, site, subsite) |>
  # keep just these columns
  dplyr::select(season_year_adj, state, region, number_functional_traps_session, number_unique_individuals_session) |>
  distinct()


```

## Overview

This report summarizes house mouse monitoring activities conducted during the 2025 survey year across Australian grain-growing regions. Mouse abundance was assessed using three complementary survey methods: live-trapping at benchmark sites, and rapid assessments using chew cards and active burrow counts.

```{r}
#| label: summary-stats
#| include: FALSE


# Calculate summary statistics for dynamic text using tbl_traps
total_individuals <- sum(tbl_traps$number_unique_individuals_session) 
total_trap_nights <- sum(tbl_traps$number_functional_traps_session) 

n_regions_traps <- data_year$traps %>%
  summarise(n = n_distinct(region)) %>%
  pull(n)

n_sites_traps <- data_year$traps %>%
  summarise(n = n_distinct(site)) %>%
  pull(n)

n_states_traps <- data_year$traps %>%
  summarise(n = n_distinct(state)) %>%
  pull(n)

n_regions_rapid <- data_year$chewcards %>%
  summarise(n = n_distinct(region)) %>%
  pull(n)

n_sites_chewcards <- data_year$chewcards %>%
  summarise(n = n_distinct(site)) %>%
  pull(n)

n_sites_burrows <- data_year$burrows %>%
  summarise(n = n_distinct(site)) %>%
  pull(n)

total_chewcards <- data_year$chewcards %>%
  summarise(total = sum(chewcards_deployed, na.rm = TRUE)) %>%
  pull(total)

chewcards_chewed <- data_year$chewcards %>%
  summarise(total = sum(chewcards_chewed, na.rm = TRUE)) %>%
  pull(total)

pct_chewed <- round((chewcards_chewed / total_chewcards) * 100, 1)

total_transects <- data_year$burrows %>%
  summarise(total = sum(burrow_transects_searched, na.rm = TRUE)) %>%
  pull(total)

transects_with_burrows <- data_year$burrows %>%
  summarise(total = sum(burrow_transects_present, na.rm = TRUE)) %>%
  pull(total)

total_active_burrows <- data_year$burrows %>%
  summarise(total = sum(burrow_total_count, na.rm = TRUE)) %>%
  pull(total)

pct_transects_burrows <- round((transects_with_burrows / total_transects) * 100, 1)

# Date ranges
date_range_traps <- data_year$traps %>%
  summarise(
    min = min(session_start_date, na.rm = TRUE),
    max = max(session_end_date, na.rm = TRUE)
  )

date_range_rapid <- bind_rows(
  data_year$chewcards %>% select(session_start_date, session_end_date),
  data_year$burrows %>% select(session_start_date, session_end_date)
) %>%
  summarise(
    min = min(session_start_date, na.rm = TRUE),
    max = max(session_end_date, na.rm = TRUE)
  )
```

**Live-trap Surveys:** A total of `r format(total_trap_nights, big.mark = ",")` trap nights (1 trap night = 1 trap set for 1 night) were conducted across `r n_sites_traps` benchmark sites in `r n_regions_traps` regions spanning `r n_states_traps` states (survey period: `r format(date_range_traps$min, "%B %Y")` to `r format(date_range_traps$max, "%B %Y")`). These surveys captured `r total_individuals` individual house mice, allowing abundance estimates at these monitoring locations.

**Rapid Assessment Surveys:** Rapid assessments were conducted across `r n_regions_rapid` regions using two methods. Chew card surveys deployed `r format(total_chewcards, big.mark = ",")` cards across `r n_sites_chewcards` sites, with `r pct_chewed`% showing evidence of mouse activity (`r chewcards_chewed` cards chewed). Active burrow surveys searched `r total_transects` transects across `r n_sites_burrows` sites (survey period: `r format(date_range_rapid$min, "%B %Y")` to `r format(date_range_rapid$max, "%B %Y")`), detecting active burrows in `r pct_transects_burrows`% of transects (total: `r total_active_burrows` active burrows).

<br>

## Live-Trap Surveys

```{r}
#| label: fig-map-traps
#| fig-height: 9
#| fig-cap: "Location of benchmark sites where house mouse were surveyed using live-traps over seasons in the current year, with colour gradient reflecting number of unique individuals caught across all subsites within the benchmark site."

# plot 
ggplot() +
  # first plot Australian states as outline in the background
  geom_sf(data = aus_shp, fill = "white") + 
  # Points colored by mice
  geom_sf(data = data_year_sf$traps, aes(color = number_unique_individuals_session), size = 3) +
  # Add viridis color scale
  scale_color_viridis_c(option = "viridis") +
  # Labels
  geom_sf_text(data = data_year_sf$traps, aes(label = region), 
               size = 2.5, nudge_y = 1.5, check_overlap = TRUE) +
  # crop map 
  coord_sf(xlim = c(128, 155), ylim = c(-40, -26), expand = FALSE) +
  # facet plot for each season 
  facet_wrap(~season_year_adj, ncol = 1) +
  # styling
  labs(color = "Individuals", y = "", x = "", subtitle = "Live-trapping") +
  theme(legend.position = "right")

```

<br>

```{r}
#| label: tbl-traps
#| tbl-cap: "Summary of live-trap survey results over seasons in the current year."

# Create a nicer formatted table (tbl_traps created in summary-stats chunk)
table_gt <- flextable(tbl_traps) %>%
   # merge cells for year and season so not repeated
  merge_v(j = "season_year_adj") |> 
  set_header_labels(
    state = "State",
    region = "Region",
    season_year_adj = "Season",
    number_functional_traps_session = "Trap Nights",
    number_unique_individuals_session = "Individuals caught"
  ) 

autofit(table_gt)

```

<br>

## Rapid Assessment Surveys

```{r}
#| label: fig-map-rapid
#| fig-width: 10
#| fig-height: 9
#| fig-cap: "Location of rapid assessment sites where house mouse were surveyed using (a) active burrow counts, and (b) chew cards in the current year. Blue dots represent sites which did not detect any mice, red dots represent sites which detected mice."


# plot active burrow counts
burrow_map = ggplot() +
  # first plot Australian states as outline in the background
  geom_sf(data = aus_shp, fill = "white") + 
  # Points colored by binary presence/absence
  geom_sf(data = data_year_sf$burrows, aes(color = factor(burrow_site_present)), size = 0.5, alpha = 0.75) +
  # Add discrete color scale
  scale_color_manual(values = c("0" = "cornflowerblue", "1" = "red"), labels = c("Absent", "Present")) +
  # crop map 
  coord_sf(xlim = c(113, 155), ylim = c(-40, -26), expand = FALSE) +
  # facet plot for each season 
  facet_wrap(~season_year_adj, ncol = 1) +
  # styling
  labs(color = "Active burrows", y = "", x = "", subtitle = "Active burrow counts") +
  theme(legend.position = "bottom")


# plot chewcards
chewcard_map = ggplot() +
  # first plot Australian states as outline in the background
  geom_sf(data = aus_shp, fill = "white") + 
  # Points colored by binary presence/absence
  geom_sf(data = data_year_sf$chewcards, aes(color = factor(chewcard_site_present)), size = 0.5, alpha = 0.75) +
  # Add discrete color scale
  scale_color_manual(values = c("0" = "cornflowerblue", "1" = "red"), labels = c("Absent", "Present")) +
  # crop map 
  coord_sf(xlim = c(113, 155), ylim = c(-40, -26), expand = FALSE) +
  # facet plot for each season 
  facet_wrap(~season_year_adj, ncol = 1) +
  # styling
  labs(color = "Chew cards", y = "", x = "", subtitle = "Chewcards") +
  theme(legend.position = "bottom")

# plot together using patchwork
 burrow_map + chewcard_map + plot_annotation(tag_levels = "a")

```

<br>

```{r}
#| label: tbl-rapid
#| tbl-cap: "Rapid assessment survey results summarised within each region. Colour gradients based on proportion of sites surveyed which detected any sign of mice (white = 0, yellow = low, orange = moderate, red = 1)."


# summary table for burrows
table_burrows <- as_tibble(data_year_sf$burrows) |>
  # summarise trapping effort per site / season
  group_by(season_year_adj, state, region) |>
  transmute(
    burrow_sites_region = n(),
    burrow_site_present_region = sum(burrow_site_present, na.rm = TRUE),
    burrow_proportion = round(burrow_site_present_region / burrow_sites_region, 2)) |>
  ungroup() |>
  distinct()

# summary table for chewcards
table_chewcards <- as_tibble(data_year_sf$chewcards) |>
  # summarise trapping effort per site / season
  group_by(season_year_adj, state, region) |>
  transmute(
    chewcard_sites_region = n(),
    chewcard_site_present_region = sum(chewcard_site_present, na.rm = TRUE),
    chewcard_proportion = round(chewcard_site_present_region / chewcard_sites_region, 2)) |>
    ungroup() |>
  distinct()

# join tables together
table_both <- full_join(table_burrows, table_chewcards,
                        by = c("state", "region", "season_year_adj")) |>
  # arrange by time first
  arrange(season_year_adj, state, region)

# Create separate color mapping functions with automatic domains
color_scale_burrows <- scales::col_numeric(
  palette = c("white", "yellow", "orange", "red"),
  domain = range(table_both$burrow_proportion, na.rm = TRUE)
)

color_scale_chewcards <- scales::col_numeric(
  palette = c("white", "yellow", "orange", "red"),
  domain = range(table_both$chewcard_proportion, na.rm = TRUE)
)




# Create a nicer formatted table
table_gt <- flextable(table_both) |>
  
  # merge cells for year and season so not repeated
  merge_v(j = c("state", "season_year_adj")) |> 

    # Color burrow columns based on burrow proportion
  bg(j = c("burrow_site_present_region", "burrow_sites_region", "burrow_proportion"),
     bg = color_scale_burrows,
     source = "burrow_proportion") |>
  
  # Color chewcard columns based on chewcard proportion
  bg(j = c("chewcard_site_present_region", "chewcard_sites_region", "chewcard_proportion"),
     bg = color_scale_chewcards,
     source = "chewcard_proportion") |>
  
  # new headings
   set_header_labels(
    state = "State",
    region = "Region",
    season_year_adj = "Season",
    burrow_sites_region = "Sites",
    burrow_site_present_region = "Presences",
    burrow_proportion = "Proportion",
    chewcard_sites_region = "Sites",
    chewcard_site_present_region = "Presences",
    chewcard_proportion = "Proportion") |>
  
   # add another header above column names
   add_header_row(values = c("", "Burrows", "Chewcards"),
                         colwidths = c(3, 3, 3), top = TRUE) |>

  # Hide the proportion columns if you don't want to display them
  #delete_columns(j = c("burrow_proportion", "chewcard_proportion")) |>
  
  # centre all values
   align(align = "center", part = "all")


table_gt

```

<br>

## Monitoring Schedule Compliance

```{r}
#| label: tbl-monitoring
#| tbl-cap: "Monitoring schedule compliance showing expected vs actual sites surveyed by region and season. Color gradients indicate performance: red = shortfall (fewer sites than expected), white = on target, green = exceeded expectations."

# Filter out rows where expected_sites = 0 and no sites were actually surveyed
# Keep rows where: expected_sites > 0 OR burrows_sites > 0 OR chewcards_sites > 0
monitoring_comparison_filtered <- monitoring_comparison |>
  dplyr::filter(expected_sites > 0 | burrows_sites > 0 | chewcards_sites > 0) |>
  # Calculate proportions for color scaling (actual / expected)
  # If expected = 0 but sites were surveyed, treat as proportion > 1 (e.g., 2)
  dplyr::mutate(
    burrows_proportion = ifelse(expected_sites == 0, 2, burrows_sites / expected_sites),
    chewcards_proportion = ifelse(expected_sites == 0, 2, chewcards_sites / expected_sites)
  ) |>
  # Arrange by season, then lead, then region to ensure proper cell merging
  dplyr::arrange(season_year_adj, lead, region) |>
  # Reorder columns: season, lead, region, expected, burrows (actual, diff), chewcards (actual, diff)
  dplyr::select(season_year_adj, lead, region, expected_sites,
         burrows_sites, burrows_difference, burrows_proportion,
         chewcards_sites, chewcards_difference, chewcards_proportion)

# Create separate color mapping functions for burrows and chewcards
# Each uses its own proportion range for independent color scaling
color_scale_burrows <- scales::col_numeric(
  palette = c("red", "white", "green"),
  domain = range(monitoring_comparison_filtered$burrows_proportion, na.rm = TRUE)
)

color_scale_chewcards <- scales::col_numeric(
  palette = c("red", "white", "green"),
  domain = range(monitoring_comparison_filtered$chewcards_proportion, na.rm = TRUE)
)

# Create flextable
monitoring_table <- flextable(monitoring_comparison_filtered) |>

  # Color burrow columns based on burrow proportion
  bg(j = c("burrows_sites", "burrows_difference", "burrows_proportion"),
     bg = color_scale_burrows,
     source = "burrows_proportion") |>

  # Color chewcard columns based on chewcard proportion
  bg(j = c("chewcards_sites", "chewcards_difference", "chewcards_proportion"),
     bg = color_scale_chewcards,
     source = "chewcards_proportion") |>

  # Set column headers (including proportion columns)
  set_header_labels(
    season_year_adj = "Season",
    lead = "Lead",
    region = "Region",
    expected_sites = "Expected",
    burrows_sites = "Actual",
    burrows_difference = "Difference",
    burrows_proportion = "Proportion",
    chewcards_sites = "Actual",
    chewcards_difference = "Difference",
    chewcards_proportion = "Proportion"
  ) |>

  # Add header row for survey types (with all 10 columns)
  add_header_row(values = c("", "", "Burrows", "Chewcards"),
                 colwidths = c(3, 1, 3, 3), top = TRUE) |>

  # Hide the proportion columns (used only for coloring)
  delete_columns(j = c("burrows_proportion", "chewcards_proportion")) |>

  # Merge cells for season and lead so not repeated
  merge_v(j = c("season_year_adj", "lead")) |>

  # Centre all values
  align(align = "center", part = "all")

monitoring_table

```
