---
title: "Rapid Assessment Data Verification Report"
subtitle: "`r basename(params$csv_path)`"
params:
  csv_path: "raw_data/survey_data/rapid_assessment_data_2026_onwards/adelaide_plains/19_01_2026.csv"
  site_info_path: null
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

# path to the CSV to verify — passed as a parameter when rendered via targets pipeline
path <- params$csv_path

# Load required packages
library(dplyr)
library(leaflet)
library(sf)
library(lubridate)
library(readr)
library(knitr)

# Load data
data <- read_csv(path, show_col_types = FALSE)

# Initialize error tracking
validation_errors <- list()
validation_passed <- TRUE
```

## Survey Information

**Data file:** `r basename(path)`
**Report generated:** `r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`
**Number of records:** `r nrow(data)`

---

# Data Validation Checks {.tabset}

```{r checks-setup}
# Expected columns and their required R class after read_csv()
# character columns are read as-is; date_in/date_out arrive as character (d/m/Y strings)
# and are parsed with dmy() later in the pipeline
expected_cols <- c(
  "project", "recorder", "organisation",
  "state", "region", "farmer", "site", "subsite", "new_site",
  "longitude", "latitude",
  "date_in", "date_out",
  "crop_type", "crop_stage", "ground_cover_percent", "biomass",
  paste0("active_burrows_t", 1:4),
  paste0("chewcard_percent_", 1:40),
  "comments"
)

# Expected class for each column (columns not listed are checked only for presence)
expected_classes <- list(
  new_site             = "logical",
  longitude            = "numeric",
  latitude             = "numeric",
  ground_cover_percent = "numeric",
  active_burrows_t1    = "numeric",
  active_burrows_t2    = "numeric",
  active_burrows_t3    = "numeric",
  active_burrows_t4    = "numeric"
)
# all chewcard columns should also be numeric
for (i in 1:40) expected_classes[[paste0("chewcard_percent_", i)]] <- "numeric"
```

## Column Structure

```{r check-columns}
missing_cols <- setdiff(expected_cols, names(data))
extra_cols <- setdiff(names(data), expected_cols)

if (length(missing_cols) > 0) {
  validation_errors <- c(validation_errors,
                         list(paste("Missing columns:", paste(missing_cols, collapse = ", "))))
  validation_passed <- FALSE
  cat("❌ **MISSING COLUMNS:**", paste(missing_cols, collapse = ", "), "\n\n")
} else {
  cat("✓ All expected columns present\n\n")
}

if (length(extra_cols) > 0) {
  cat("⚠️  **UNEXPECTED COLUMNS:**", paste(extra_cols, collapse = ", "), "\n\n")
}
```

## Blank Cells

```{r check-blanks}
blank_issues <- data %>%
  mutate(across(everything(), as.character)) %>%
  summarise(across(everything(), ~sum(.x == "", na.rm = TRUE))) %>%
  select(where(~.x > 0))

if (ncol(blank_issues) > 0) {
  validation_errors <- c(validation_errors,
                         list("Blank cells found (should be explicit NA)"))
  validation_passed <- FALSE
  cat("❌ **BLANK CELLS FOUND** (should be NA):\n\n")
  kable(blank_issues)
} else {
  cat("✓ No blank cells found\n\n")
}
```

## Required Fields

```{r check-required}
required_fields <- c("project", "recorder", "state", "region",
                     "farmer", "site",
                     "new_site", "longitude", "latitude", "date_in", "date_out")

na_counts <- data %>%
  select(all_of(required_fields[required_fields %in% names(data)])) %>%
  summarise(across(everything(), ~sum(is.na(.x)))) %>%
  select(where(~.x > 0))

if (ncol(na_counts) > 0) {
  validation_errors <- c(validation_errors,
                         list("Missing values in required fields"))
  validation_passed <- FALSE
  cat("❌ **MISSING VALUES IN REQUIRED FIELDS:**\n\n")
  kable(na_counts)
} else {
  cat("✓ All required fields have values\n\n")
}
```

## Coordinates

```{r check-coords}
coord_issues <- character()

## 1. Range checks — longitude 110-155, latitude -45 to -10
if ("longitude" %in% names(data)) {
  invalid_lon <- sum(data$longitude < 110 | data$longitude > 155, na.rm = TRUE)
  if (invalid_lon > 0) {
    coord_issues <- c(coord_issues, paste(invalid_lon, "longitude value(s) outside Australian range (110-155)"))
    validation_errors <- c(validation_errors, list("Invalid longitude values"))
    validation_passed <- FALSE
    cat("❌", invalid_lon, "longitude value(s) outside Australian range (110–155)\n\n")
  } else {
    cat("✓ Longitude values within Australian range (110–155)\n\n")
  }
}

if ("latitude" %in% names(data)) {
  invalid_lat <- sum(data$latitude < -45 | data$latitude > -10, na.rm = TRUE)
  if (invalid_lat > 0) {
    coord_issues <- c(coord_issues, paste(invalid_lat, "latitude value(s) outside Australian range (-45 to -10)"))
    validation_errors <- c(validation_errors, list("Invalid latitude values"))
    validation_passed <- FALSE
    cat("❌", invalid_lat, "latitude value(s) outside Australian range (−45 to −10)\n\n")
  } else {
    cat("✓ Latitude values within Australian range (−45 to −10)\n\n")
  }
}

## 2. Swapped coordinate check — catch the common mistake of entering (lat, lon) instead of (lon, lat)
if (all(c("longitude", "latitude") %in% names(data))) {
  # longitude should be large positive (~110-155); latitude should be negative (~-45 to -10)
  # swapped if longitude looks like a latitude (negative or tiny) or latitude looks like a longitude (large positive)
  swapped <- with(data, (longitude < 0 | longitude < 90) & (latitude > 0 | latitude > 100))
  n_swapped <- sum(swapped, na.rm = TRUE)
  if (n_swapped > 0) {
    coord_issues <- c(coord_issues, paste(n_swapped, "record(s) with possibly swapped lat/lon"))
    validation_errors <- c(validation_errors, list("Possibly swapped latitude and longitude"))
    validation_passed <- FALSE
    cat("❌", n_swapped, "record(s) where latitude and longitude may be swapped\n\n")
    kable(data[which(swapped), c("farmer", "site", "subsite", "state", "region", "longitude", "latitude")],
          caption = "Possibly swapped coordinates")
  } else {
    cat("✓ No obviously swapped lat/lon detected\n\n")
  }
}

## 3. Spatial validation — check points fall within Australian state boundaries and
##    that the stated 'state' field matches the state polygon the point falls in.
##    Uses aus_shp from the targets store (ST_NAME column = state abbreviation, e.g. "NSW").
if (all(c("longitude", "latitude") %in% names(data))) {
  tryCatch({
    store_path <- if (dir.exists("_targets")) "_targets" else "../_targets"
    aus_shp <- targets::tar_read(aus_shp, store = store_path)

    # only check rows that have valid coordinates
    data_coords <- data[!is.na(data$longitude) & !is.na(data$latitude), ]
    data_sf <- sf::st_as_sf(data_coords, coords = c("longitude", "latitude"), crs = 4326)

    # dissolve state polygons into a single Australia boundary
    aus_boundary <- sf::st_union(aus_shp)

    # check every point is inside Australia
    within_aus <- as.vector(sf::st_within(data_sf, aus_boundary, sparse = FALSE))
    n_outside <- sum(!within_aus)
    if (n_outside > 0) {
      coord_issues <- c(coord_issues, paste(n_outside, "point(s) outside Australian boundaries"))
      validation_errors <- c(validation_errors, list("Point(s) fall outside Australia"))
      validation_passed <- FALSE
      cat("❌", n_outside, "point(s) fall outside Australian state boundaries\n\n")
      kable(data_coords[!within_aus, c("farmer", "site", "subsite", "state", "region", "longitude", "latitude")],
            caption = "Records outside Australia")
    } else {
      cat("✓ All points fall within Australian state boundaries\n\n")
    }

    # check stated state matches the state polygon the point falls in
    # aus_shp$Name holds the full state name (e.g. "Western Australia") which matches
    # the CSV convention; ST_NAME holds the abbreviation (e.g. "WA") for display
    if ("state" %in% names(data)) {
      joined <- sf::st_join(data_sf, aus_shp[, c("ST_NAME", "Name")], join = sf::st_within)
      joined_df <- sf::st_drop_geometry(joined)
      # compare case-insensitively using full Name (CSV has e.g. "western australia")
      state_mismatch <- !is.na(joined_df$Name) &
        tolower(trimws(data_coords$state)) != tolower(trimws(joined_df$Name))
      n_mismatch <- sum(state_mismatch, na.rm = TRUE)
      if (n_mismatch > 0) {
        coord_issues <- c(coord_issues, paste(n_mismatch, "record(s) with state field not matching coordinate location"))
        validation_errors <- c(validation_errors, list("State field does not match coordinate location"))
        validation_passed <- FALSE
        cat("❌", n_mismatch, "record(s) where stated 'state' does not match the state the coordinates fall in\n\n")
        mismatch_rows <- cbind(
          data_coords[state_mismatch, c("farmer", "site", "subsite", "state", "region", "longitude", "latitude")],
          state_from_coords = paste0(joined_df$Name[state_mismatch], " (", joined_df$ST_NAME[state_mismatch], ")")
        )
        kable(mismatch_rows, caption = "State field vs. coordinate-derived state")
      } else {
        cat("✓ All stated 'state' values match the coordinates' location\n\n")
      }
    }

  }, error = function(e) {
    cat("⚠️ Spatial validation could not be completed:", conditionMessage(e), "\n\n")
  })
}

## 4. Coordinate table — display all records so values can be visually inspected
if (all(c("farmer", "site", "subsite", "state", "region", "longitude", "latitude") %in% names(data))) {
  cat("\n**All site coordinates (for visual inspection):**\n\n")
  kable(data[, c("farmer", "site", "subsite", "state", "region", "longitude", "latitude")],
        caption = "All site coordinates")
}
```

```{r map-coords}
#| fig-width: 10
#| fig-height: 6

# only render if coordinates are present
if (all(c("longitude", "latitude") %in% names(data)) &&
    any(!is.na(data$longitude) & !is.na(data$latitude))) {

  # popup shows key fields per record
  popup_text <- paste0(
    "<b>", data$farmer, " – ", data$site, "</b><br>",
    "Subsite: ", data$subsite, "<br>",
    "Region: ", data$region, "<br>",
    "State: ", data$state, "<br>",
    "Lon: ", round(data$longitude, 6), "  Lat: ", round(data$latitude, 6), "<br>",
    "New site: ", data$new_site
  )

  # colour markers by region so multiple regions are easy to distinguish
  unique_regions <- unique(data$region)
  region_colors  <- rainbow(length(unique_regions))
  names(region_colors) <- unique_regions
  marker_colors <- region_colors[data$region]

  leaflet(data) |>
    addProviderTiles(provider = "Esri.WorldImagery", group = "Satellite") |>
    addProviderTiles(provider = "OpenStreetMap",     group = "Street Map") |>
    addCircleMarkers(
      lng         = ~longitude,
      lat         = ~latitude,
      popup       = popup_text,
      label       = ~paste(farmer, site, subsite, sep = " / "),
      color       = marker_colors,
      fillColor   = marker_colors,
      radius      = 8,
      fillOpacity = 0.8,
      stroke      = TRUE,
      weight      = 2
    ) |>
    addScaleBar(position = "bottomleft", options = scaleBarOptions(imperial = FALSE)) |>
    addLayersControl(
      baseGroups = c("Street Map", "Satellite"),
      options    = layersControlOptions(collapsed = FALSE)
    ) |>
    addLegend(
      position = "bottomright",
      colors   = region_colors,
      labels   = names(region_colors),
      title    = "Region",
      opacity  = 0.8
    )
}
```

## Existing Sites

```{r check-existing-sites}
# For rows marked new_site == FALSE, verify that farmer/site/subsite already
# exists in data_site_information (the reference table built from all historical
# survey data). Also check that coordinates are close to the recorded location
# (within 0.05 degrees, ~5 km) to catch accidental coordinate edits.

if ("new_site" %in% names(data) && all(c("region", "farmer", "site", "subsite") %in% names(data))) {

  tryCatch({
    # prefer the RDS passed in via params (set by the targets pipeline to avoid
    # calling tar_read() mid-run which causes metadata race conditions);
    # fall back to tar_read() when rendering the QMD manually outside the pipeline
    if (!is.null(params$site_info_path) && file.exists(params$site_info_path)) {
      site_info <- readRDS(params$site_info_path)
    } else {
      store_path <- if (dir.exists("_targets")) "_targets" else "../_targets"
      site_info  <- targets::tar_read(data_site_information, store = store_path)
    }

    # only check rows marked as existing sites
    existing <- data[!is.na(data$new_site) & data$new_site == FALSE, ]

    if (nrow(existing) == 0) {
      cat("ℹ️ No rows marked as existing sites (new_site == FALSE)\n\n")
    } else {

      ## 1. Check that the region/farmer/site/subsite combination is recognised.
      ## All four columns must match exactly (case-insensitive) against the
      ## site database — a mismatch in any one field is flagged as unrecognised.
      existing_key <- tolower(trimws(paste(existing$region, existing$farmer,
                                           existing$site,   existing$subsite)))
      known_key    <- tolower(trimws(paste(site_info$region, site_info$farmer,
                                           site_info$site,   site_info$subsite)))
      unrecognised <- !existing_key %in% known_key

      if (any(unrecognised)) {
        validation_errors <- c(validation_errors,
          list(paste(sum(unrecognised),
               "existing site(s) not found in site information database")))
        validation_passed <- FALSE
        cat("❌", sum(unrecognised),
            "row(s) marked new_site = FALSE but region/farmer/site/subsite",
            "not found in the site database\n\n")
        kable(existing[unrecognised,
                       c("region", "farmer", "site", "subsite", "state", "longitude", "latitude")],
              caption = "Unrecognised existing sites (region/farmer/site/subsite not matched)")
      } else {
        cat("✓ All existing sites (new_site = FALSE) have a matching region/farmer/site/subsite in the site database\n\n")
      }

      ## 2. For recognised sites, check that coordinates are close to the stored values.
      ## Uses the same four-column key so the correct stored record is matched.
      if (all(c("longitude", "latitude") %in% names(existing))) {
        recognised <- existing[!unrecognised, ]

        if (nrow(recognised) > 0) {
          recognised_key <- tolower(trimws(paste(recognised$region, recognised$farmer,
                                                 recognised$site,   recognised$subsite)))
          matched_info   <- site_info[match(recognised_key, known_key), ]

          lon_diff <- abs(recognised$longitude - matched_info$longitude)
          lat_diff <- abs(recognised$latitude  - matched_info$latitude)
          # flag if either dimension differs by more than 0.05 degrees (~5 km)
          coord_changed <- (lon_diff > 0.05 | lat_diff > 0.05) & !is.na(lon_diff) & !is.na(lat_diff)

          if (any(coord_changed)) {
            cat("⚠️", sum(coord_changed),
                "existing site(s) have coordinates that differ from stored values by > 0.05°\n\n")
            coord_check_rows <- cbind(
              recognised[coord_changed, c("region", "farmer", "site", "subsite", "longitude", "latitude")],
              stored_longitude = round(matched_info$longitude[coord_changed], 6),
              stored_latitude  = round(matched_info$latitude[coord_changed],  6),
              lon_diff         = round(lon_diff[coord_changed], 6),
              lat_diff         = round(lat_diff[coord_changed], 6)
            )
            kable(coord_check_rows, caption = "Existing sites with coordinates differing from stored values")
          } else {
            cat("✓ Coordinates for all existing sites are consistent with stored values\n\n")
          }
        }
      }
    }

  }, error = function(e) {
    cat("⚠️ Existing site check could not be completed:", conditionMessage(e), "\n\n")
  })

} else {
  cat("ℹ️ Skipped: new_site, region, farmer, site, or subsite column not present\n\n")
}
```

## Ground Cover

```{r check-ground-cover}
if ("ground_cover_percent" %in% names(data)) {
  invalid_gc <- sum(data$ground_cover_percent < 0 | data$ground_cover_percent > 100, na.rm = TRUE)
  if (invalid_gc > 0) {
    validation_errors <- c(validation_errors, list("Invalid ground cover percentages"))
    validation_passed <- FALSE
    cat("❌", invalid_gc, "ground_cover_percent value(s) outside 0-100 range\n\n")
  } else {
    cat("✓ Ground cover percentage values valid\n\n")
  }
}
```

## Burrow Counts

```{r check-burrows}
burrow_cols <- paste0("active_burrows_t", 1:4)
burrow_cols <- burrow_cols[burrow_cols %in% names(data)]

if (length(burrow_cols) > 0) {
  invalid_burrows <- data %>%
    select(all_of(burrow_cols)) %>%
    summarise(across(everything(), ~sum(.x < 0, na.rm = TRUE))) %>%
    select(where(~.x > 0))

  if (ncol(invalid_burrows) > 0) {
    validation_errors <- c(validation_errors, list("Negative burrow counts"))
    validation_passed <- FALSE
    cat("❌ **NEGATIVE VALUES IN BURROW COUNTS:**\n\n")
    kable(invalid_burrows)
  } else {
    cat("✓ All burrow counts are ≥ 0\n\n")
  }
}
```

## Chewcard Percentages

```{r check-chewcards}
chewcard_cols <- paste0("chewcard_percent_", 1:40)
chewcard_cols <- chewcard_cols[chewcard_cols %in% names(data)]

if (length(chewcard_cols) > 0) {
  invalid_chew <- data %>%
    select(all_of(chewcard_cols)) %>%
    summarise(across(everything(), ~sum(.x < 0 | .x > 100, na.rm = TRUE))) %>%
    select(where(~.x > 0))

  if (ncol(invalid_chew) > 0) {
    validation_errors <- c(validation_errors, list("Invalid chewcard percentages"))
    validation_passed <- FALSE
    cat("❌ **INVALID CHEWCARD PERCENTAGES** (should be 0-100):\n\n")
    kable(invalid_chew)
  } else {
    cat("✓ All chewcard percentages are 0-100\n\n")
  }
}
```

## Dates

```{r check-dates}
if ("date_in" %in% names(data) && "date_out" %in% names(data)) {
  date_in_parsed  <- dmy(data$date_in)
  date_out_parsed <- dmy(data$date_out)

  # --- 1. date_out before date_in ---
  date_issues <- sum(date_out_parsed < date_in_parsed, na.rm = TRUE)
  if (date_issues > 0) {
    validation_errors <- c(validation_errors, list("Invalid date ranges"))
    validation_passed <- FALSE
    cat("❌", date_issues, "record(s) where date_out is before date_in\n\n")
  } else {
    cat("✓ Date ranges are logical (date_out ≥ date_in for all records)\n\n")
  }

  # --- 2. Survey duration > 6 days ---
  # A chewcard session longer than 6 days is highly unusual for rapid assessment
  # surveys and most likely indicates a date entry error (e.g. wrong month or year).
  duration_days <- as.numeric(date_out_parsed - date_in_parsed)
  long_survey   <- !is.na(duration_days) & duration_days > 6

  if (any(long_survey)) {
    validation_errors <- c(validation_errors,
      list(paste(sum(long_survey), "record(s) with survey duration > 6 days")))
    validation_passed <- FALSE
    cat("❌", sum(long_survey), "record(s) with a survey duration longer than 6 days\n\n")
    long_rows <- data[long_survey,
                      intersect(c("farmer", "site", "subsite", "date_in", "date_out"), names(data))]
    long_rows$duration_days <- duration_days[long_survey]
    kable(long_rows, caption = "Records with unusually long survey duration (> 6 days)")
  } else {
    cat("✓ All survey durations are ≤ 6 days\n\n")
  }

  # --- 3. First survey night ≠ session start date ---
  # date_in is the session start date (the night chewcards are deployed) and therefore
  # represents the first — and for rapid assessment, only — survey night.  If date_in
  # equals date_out the cards were collected the same day they were set out, meaning
  # there was no overnight soak period.  This is almost certainly a data entry error.
  same_day <- !is.na(date_in_parsed) & !is.na(date_out_parsed) &
              date_in_parsed == date_out_parsed

  if (any(same_day)) {
    validation_errors <- c(validation_errors,
      list(paste(sum(same_day),
           "record(s) where date_in equals date_out (first survey night = session end, no overnight soak)")))
    validation_passed <- FALSE
    cat("❌", sum(same_day),
        "record(s) where date_in == date_out",
        "(first survey night equals the session end date — likely entry error)\n\n")
    kable(data[same_day,
               intersect(c("farmer", "site", "subsite", "date_in", "date_out"), names(data))],
          caption = "Records where date_in equals date_out (no overnight soak)")
  } else {
    cat("✓ All records have date_out after date_in (overnight soak period present)\n\n")
  }

  # --- 4. Year outlier check ---
  # Detects obvious year typos, e.g. one record shows 2025 while all others are 2026.
  # Checks both date_in and date_out years against the dominant (most frequent) year
  # across all dates in the file.
  all_years <- year(c(date_in_parsed, date_out_parsed))
  all_years <- all_years[!is.na(all_years)]

  if (length(all_years) > 0) {
    dominant_year  <- as.integer(names(sort(table(all_years), decreasing = TRUE)[1]))
    outlier_in     <- !is.na(date_in_parsed)  & year(date_in_parsed)  != dominant_year
    outlier_out    <- !is.na(date_out_parsed) & year(date_out_parsed) != dominant_year
    has_year_outlier <- outlier_in | outlier_out

    if (any(has_year_outlier)) {
      validation_errors <- c(validation_errors,
        list(paste0(sum(has_year_outlier),
             " record(s) with a date year that differs from the dominant year (",
             dominant_year, ")")))
      validation_passed <- FALSE
      cat("❌", sum(has_year_outlier),
          "record(s) with a date year different from the dominant year (", dominant_year, ")\n\n")
      outlier_rows <- data[has_year_outlier,
                           intersect(c("farmer", "site", "subsite", "date_in", "date_out"), names(data))]
      kable(outlier_rows, caption = paste("Records with year ≠", dominant_year,
                                          "(dominant year across all dates in this file)"))
    } else {
      cat("✓ All date years are consistent with the dominant year (", dominant_year, ")\n\n")
    }
  }
}
```

## Biomass Values

```{r check-biomass}
if ("biomass" %in% names(data)) {
  valid_biomass <- c("low", "medium", "high", NA)
  invalid_biomass <- sum(!tolower(data$biomass) %in% valid_biomass, na.rm = TRUE)
  if (invalid_biomass > 0) {
    validation_errors <- c(validation_errors, list("Invalid biomass values"))
    validation_passed <- FALSE
    cat("❌", invalid_biomass, "biomass value(s) not in (low, medium, high)\n\n")
    cat("Unique values found:", paste(unique(data$biomass), collapse = ", "), "\n\n")
  } else {
    cat("✓ Biomass values valid\n\n")
  }
}
```

## Data Types

```{r check-types}
# check that each column present in the data matches its expected R class
type_mismatches <- data.frame(
  column   = character(),
  expected = character(),
  actual   = character(),
  stringsAsFactors = FALSE
)

for (col in names(expected_classes)) {
  if (col %in% names(data)) {
    actual_class <- class(data[[col]])[1]
    exp_class    <- expected_classes[[col]]
    # treat integer as numeric (readr sometimes reads whole-number columns as integer)
    # all-NA columns are read as logical by readr regardless of intended type;
    # this is harmless because bind_rows() will coerce correctly when combined
    # with other files that have real values in that column
    all_na <- all(is.na(data[[col]]))
    ok <- actual_class == exp_class ||
          (exp_class == "numeric" && actual_class == "integer") ||
          (exp_class == "numeric" && actual_class == "logical" && all_na)
    if (!ok) {
      type_mismatches <- rbind(type_mismatches, data.frame(
        column   = col,
        expected = exp_class,
        actual   = actual_class,
        stringsAsFactors = FALSE
      ))
    }
  }
}

if (nrow(type_mismatches) > 0) {
  validation_errors <- c(validation_errors, list("Column(s) have incorrect data type"))
  validation_passed <- FALSE
  cat("❌ Column type mismatches found:\n\n")
  kable(type_mismatches, col.names = c("Column", "Expected class", "Actual class"))
} else {
  cat("✓ All columns have the correct data type\n\n")
}
```

---

```{r halt-if-errors}
# Stop rendering if validation failed
if (!validation_passed) {
  cat("\n\n# ⚠️ VALIDATION FAILED\n\n")
  cat("**The following errors were found:**\n\n")
  for (i in seq_along(validation_errors)) {
    cat(i, ". ", validation_errors[[i]], "\n", sep = "")
  }
  cat("\n\n**Please correct these issues before proceeding.**\n\n")
  knitr::knit_exit()
}
```

::: {.callout-note appearance="simple"}
## Validation Status
✅ **All validation checks passed!** Proceeding with summary and analysis.
:::

---

# Data Summary

```{r summary-survey}
cat("## Survey Details\n\n")
cat("- **Project(s):** ", paste(unique(data$project), collapse = ", "), "\n")
cat("- **Recorder(s):** ", paste(unique(data$recorder), collapse = ", "), "\n")
cat("- **Organisation(s):** ", paste(unique(data$organisation), collapse = ", "), "\n")
```

```{r summary-sites}
cat("\n## Site Information\n\n")
cat("- **Total records:** ", nrow(data), "\n")
cat("- **Unique farmers:** ", length(unique(data$farmer)), "\n")
cat("- **Unique sites:** ", length(unique(paste(data$farmer, data$site))), "\n")
cat("- **Unique subsites:** ", length(unique(paste(data$farmer, data$site, data$subsite))), "\n")

if ("new_site" %in% names(data)) {
  cat("- **New sites:** ", sum(data$new_site, na.rm = TRUE), "\n")
  cat("- **Established sites:** ", sum(!data$new_site, na.rm = TRUE), "\n")
}

cat("- **State(s):** ", paste(unique(data$state), collapse = ", "), "\n")
cat("- **Region(s):** ", paste(unique(data$region), collapse = ", "), "\n")
```

```{r summary-dates}
cat("\n## Survey Period\n\n")
if ("date_in" %in% names(data) && "date_out" %in% names(data)) {
  date_in_parsed <- dmy(data$date_in)
  date_out_parsed <- dmy(data$date_out)

  cat("- **Survey start dates:** ", paste(range(date_in_parsed, na.rm = TRUE), collapse = " to "), "\n")
  cat("- **Survey end dates:** ", paste(range(date_out_parsed, na.rm = TRUE), collapse = " to "), "\n")

  survey_days <- as.numeric(date_out_parsed - date_in_parsed)
  cat("- **Survey duration (days):** min = ", min(survey_days, na.rm = TRUE),
      ", max = ", max(survey_days, na.rm = TRUE),
      ", mean = ", round(mean(survey_days, na.rm = TRUE), 1), "\n", sep = "")
}
```

```{r summary-crop}
cat("\n## Crop Information\n\n")
if ("crop_type" %in% names(data)) {
  cat("- **Crop type(s):** ", paste(unique(data$crop_type), collapse = ", "), "\n")
}
if ("crop_stage" %in% names(data)) {
  cat("- **Crop stage(s):** ", paste(unique(data$crop_stage), collapse = ", "), "\n")
}
if ("biomass" %in% names(data)) {
  biomass_table <- table(data$biomass)
  cat("- **Biomass levels:** ", paste(names(biomass_table), "=", biomass_table, collapse = ", "), "\n")
}
```

```{r summary-burrows}
cat("\n## Active Burrow Transects\n\n")
burrow_cols <- paste0("active_burrows_t", 1:4)
burrow_cols <- burrow_cols[burrow_cols %in% names(data)]

if (length(burrow_cols) > 0) {
  burrow_data <- data %>% select(all_of(burrow_cols))

  transects_surveyed <- rowSums(!is.na(burrow_data))
  cat("- **Transects per site:** min = ", min(transects_surveyed),
      ", max = ", max(transects_surveyed),
      ", mean = ", round(mean(transects_surveyed), 1), "\n", sep = "")

  total_burrows <- rowSums(burrow_data, na.rm = TRUE)
  cat("- **Total active burrows per site:**\n")
  cat("  - Range: ", min(total_burrows), " - ", max(total_burrows), "\n", sep = "")
  cat("  - Mean: ", round(mean(total_burrows), 1), "\n", sep = "")
  cat("  - Median: ", median(total_burrows), "\n", sep = "")
  cat("- **Sites with 0 burrows:** ", sum(total_burrows == 0), "\n", sep = "")
}
```

```{r summary-chewcards}
cat("\n## Chewcard Results\n\n")
chewcard_cols <- paste0("chewcard_percent_", 1:40)
chewcard_cols <- chewcard_cols[chewcard_cols %in% names(data)]

if (length(chewcard_cols) > 0) {
  chewcard_data <- data %>% select(all_of(chewcard_cols))

  cards_deployed <- rowSums(!is.na(chewcard_data))
  cat("- **Chewcards deployed per site:** min = ", min(cards_deployed),
      ", max = ", max(cards_deployed),
      ", mean = ", round(mean(cards_deployed), 1), "\n", sep = "")

  cards_chewed <- rowSums(chewcard_data > 0, na.rm = TRUE)
  cat("- **Cards with chewing per site:**\n")
  cat("  - Range: ", min(cards_chewed), " - ", max(cards_chewed), "\n", sep = "")
  cat("  - Mean: ", round(mean(cards_chewed), 1), "\n", sep = "")

  all_percentages <- unlist(chewcard_data)
  all_percentages <- all_percentages[!is.na(all_percentages)]

  if (length(all_percentages) > 0) {
    cat("- **Chewing percentage across all cards:**\n")
    cat("  - Range: ", min(all_percentages), " - ", max(all_percentages), "%\n", sep = "")
    cat("  - Mean: ", round(mean(all_percentages), 1), "%\n", sep = "")
    cat("  - Median: ", median(all_percentages), "%\n", sep = "")
    cat("  - Cards with >0% chewing: ", sum(all_percentages > 0),
        " (", round(100 * sum(all_percentages > 0) / length(all_percentages), 1), "%)\n", sep = "")
  }
}
```

