---
title: "Sites sharing a paddock"
date: last-modified
format:
  html:
    self-contained: true
    theme: cosmo
    toc: true
execute:
  echo: false
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup
#| include: false

library(targets)
library(dplyr)
library(sf)
library(leaflet)
library(stringr)
library(flextable)
library(glue)

store_path <- if (dir.exists("_targets")) "_targets" else "../../_targets"
targets::tar_load(c(duplicate_paddock_sites, duplicate_paddock_polygons),
                  store = store_path)

# Classify likely cause of sharing to guide investigation
duplicate_paddock_sites <- duplicate_paddock_sites |>
  dplyr::group_by(paddock_id) |>
  dplyr::mutate(
    likely_cause = dplyr::case_when(
      # Any site was snapped rather than falling within — GPS outside boundary
      any(!is.na(paddock_snap_dist_m)) ~ "GPS outside polygon (snapped)",
      # Region strings differ only by space vs underscore → naming variant
      dplyr::n_distinct(gsub("_", " ", region)) == 1 &
        dplyr::n_distinct(region) > 1 ~ "Region name variant",
      # Same farmer and site, different subsites → multiple transects per paddock
      dplyr::n_distinct(farmer) == 1 &
        dplyr::n_distinct(site) == 1 ~ "Multiple subsites in paddock",
      # Same farmer but different site names
      dplyr::n_distinct(farmer) == 1 ~ "Same farmer, different site names",
      # Different farmers sharing a paddock
      TRUE ~ "Different farmers in same paddock"
    )
  ) |>
  dplyr::ungroup()

cause_order <- c(
  "Region name variant",
  "GPS outside polygon (snapped)",
  "Multiple subsites in paddock",
  "Same farmer, different site names",
  "Different farmers in same paddock"
)
duplicate_paddock_sites$likely_cause <- factor(
  duplicate_paddock_sites$likely_cause,
  levels = cause_order
)
```

## Overview

**`r dplyr::n_distinct(duplicate_paddock_sites$paddock_id)`** paddocks are each
associated with more than one recorded site
(**`r nrow(duplicate_paddock_sites)`** site rows total across
**`r dplyr::n_distinct(duplicate_paddock_sites$state)`** states).

```{r}
#| label: summary-table

duplicate_paddock_sites |>
  dplyr::count(likely_cause, name = "n_sites") |>
  dplyr::arrange(likely_cause) |>
  flextable::flextable() |>
  flextable::set_header_labels(
    likely_cause = "Likely cause",
    n_sites      = "Sites"
  ) |>
  flextable::autofit() |>
  flextable::theme_vanilla()
```

## Map

Each paddock polygon is coloured by its likely cause. Click a polygon or marker
for details.

```{r}
#| label: map

# One colour per cause category
cause_pal <- leaflet::colorFactor(
  palette = c("#E69F00", "#D55E00", "#009E73", "#0072B2", "#CC79A7"),
  domain  = cause_order
)

# Join cause onto polygon layer via paddock id
poly_with_cause <- duplicate_paddock_polygons |>
  dplyr::left_join(
    dplyr::distinct(duplicate_paddock_sites, paddock_id, likely_cause),
    by = c("id" = "paddock_id")
  )

# Site-level popups
site_popups <- glue::glue_data(duplicate_paddock_sites,
  "<b>Paddock ID:</b> {paddock_id}<br>
   <b>{str_to_title(site)}</b> — {str_to_title(subsite)}<br>
   <b>Farmer:</b> {str_to_title(farmer)}<br>
   <b>Region:</b> {str_to_title(region)}, {state}<br>
   <b>Coords:</b> {round(longitude, 4)}, {round(latitude, 4)}<br>
   <b>Cause:</b> {likely_cause}"
)

leaflet::leaflet() |>
  leaflet::addProviderTiles(leaflet::providers$Esri.WorldImagery) |>
  leaflet::addPolygons(
    data        = poly_with_cause,
    color       = "#333333",
    weight      = 1.5,
    opacity     = 0.9,
    fillColor   = ~cause_pal(likely_cause),
    fillOpacity = 0.45,
    label       = ~as.character(id),
    group       = "Paddocks"
  ) |>
  leaflet::addCircleMarkers(
    data        = duplicate_paddock_sites,
    lng         = ~longitude,
    lat         = ~latitude,
    radius      = 6,
    color       = "white",
    weight      = 2,
    fillColor   = ~cause_pal(likely_cause),
    fillOpacity = 0.95,
    popup       = site_popups,
    label       = ~str_to_title(paste0(site, " / ", subsite)),
    group       = "Sites"
  ) |>
  leaflet::addLegend(
    position  = "bottomright",
    pal       = cause_pal,
    values    = cause_order,
    title     = "Likely cause",
    na.label  = ""
  ) |>
  leaflet::addLayersControl(
    overlayGroups = c("Paddocks", "Sites"),
    options       = leaflet::layersControlOptions(collapsed = FALSE)
  )
```

## Site details by paddock

Each group is one paddock. The **likely cause** column is a heuristic —
review each case to determine the appropriate action (merge site names, correct
region spelling, or accept as genuine distinct subsites).

```{r}
#| label: details-table

duplicate_paddock_sites |>
  dplyr::mutate(
    snap_dist = dplyr::if_else(
      !is.na(paddock_snap_dist_m),
      paste0(paddock_snap_dist_m, " m"),
      "—"
    ),
    longitude = round(longitude, 4),
    latitude  = round(latitude,  4)
  ) |>
  dplyr::select(
    paddock_id, likely_cause, state, region, farmer,
    site, subsite, longitude, latitude, snap_dist
  ) |>
  flextable::as_grouped_data(groups = "paddock_id") |>
  flextable::flextable() |>
  flextable::set_header_labels(
    paddock_id  = "Paddock ID",
    likely_cause = "Likely cause",
    state       = "State",
    region      = "Region",
    farmer      = "Farmer",
    site        = "Site",
    subsite     = "Subsite",
    longitude   = "Lon",
    latitude    = "Lat",
    snap_dist   = "Snap dist"
  ) |>
  flextable::autofit() |>
  flextable::theme_vanilla() |>
  flextable::fontsize(size = 9, part = "all")
```
